A neuronal least-action principle for real-time learning in cortical circuits

Version 1

Abstract

Abstract

One of the most fundamental laws of physics is the principle of least action. Motivated by its predictive power, we introduce a neural least-action principle that we apply to motor control. The central notion is the somato-dendritic mismatch error within individual neurons. The principle postulates that the somato-dendritic mismatch errors across all neurons in a cortical network are minimized by the voltage dynamics. Ongoing synaptic plasticity reduces the somato-dendritic mismatch error within each neuron and performs gradient descent on the output cost in real time. The neuronal activity is prospective, ensuring that dendritic errors deep in the network are prospectively corrected to eventually reduce motor errors. The neuron-specific errors are represented in the apical dendrites of pyramidal neurons, and are extracted by a cortical microcircuit that ‘explains away’ the feedback from the periphery. The principle offers a general theoretical framework to functionally describe real-time neuronal and synaptic processing.

Introduction

Wigner’s remark about the ‘unreasonable effectiveness’ of mathematics in allowing us to understand physical phenomena (Wigner, 1959) is famously contrasted by Gelfand’s quip about its ‘unreasonable ineffectiveness’ in doing the same for biology (Borovik, 2021). Considering the component of randomness that is inherent to evolution, this may not be all that surprising. However, while this argument holds just as well for the brain at the cellular level, ultimately brains are computing devices. At the level of computation, machine learning and neuroscience have revealed near-optimal strategies for information processing and storage, and evolution is likely to have found similar principles through trial and error (Hassabis et al., 2017). Thus, we have reason to hope for the existence of fundamental principles of cortical computation that are similar to those we have found in the physical sciences. Eventually, it is important for such approaches to relate these principles back to brain phenomenology and connect function to structure and dynamics.

In physics, a fundamental measure of ‘effort’ is the action of a system, which nature seeks to ‘minimize’. Given an appropriate description of interactions between the systems constituents, the least-action principle can be used to derive the equations of motion of any physical system (Feynman et al., 2011; Coopersmith, 2017). Here, we suggest that in biological information processing, a similar principle holds for prediction errors, which are of obvious relevance for cognition and behavior. Based on such errors, we formulate a neuronal least-action (NLA) principle which can be used to derive neuronal dynamics and map them to observed dendritic morphologies and cortical microcircuits. Within this framework, local synaptic plasticity at basal and apical dendrites can be derived by stochastic gradient descent on errors. The errors that are minimized refer to the errors in output neurons that are typically thought to represent motor trajectories, planned and encoded in cortical motor areas and ultimately in the spinal cord and muscles. In the context of motor control, a phenomenological minimal action principle has previously been proposed that guides the planning and execution of movements (Feldman & Levin, 2009). The present paper formalizes this principle and reformulates the classical equilibrium point hypothesis (Latash, 2010) in a dynamical setting. Other attempts exist to link neural networks with the least action principle, for instance by directly learning to reproduce a given trajectory (Amirikian & Lukashin, 1992), by minimizing data transport through a network (Karkar et al., 2021), or by minimizing the free energy (Friston, 2010; Friston et al., 2022).

Our theory considers the minimization of a mismatch between the effective output trajectory and an internally or externally imposed target trajectory. The strength by which the output trajectory is pushed towards the target trajectory is referred to as ‘nudging strength’. For weak nudging, the feedback slightly adapts the internal network states so that the output activities closer follow their targets. For a target that is fixed in time, the dynamics reduces to the Equilibrium Propagation (Scellier & Bengio, 2017). For strong nudging, the output neurons are clamped to tightly follow the target trajectory (Song et al., 2022; Meulemans, Zucchet, et al., 2022). Each network neuron performs gradient descent on its own dendritic prediction error (Urbanczik & Senn, 2014), and with this eventually also contributes to a reduction of the output errors. For both versions, weak and strong nudging, synaptic plasticity in an interneuron feedback circuitry supports the extraction of neuron-specific dendritic errors (Sacramento et al., 2018), providing a cortical implementation of the suggested ‘real-time dendritic error propagation (rt-DeEP)’.

The NLA principle can be put into the history of energy-based models used to understand neuronal processing and learning in recurrent neural networks (Hopfield, 1982), artificial intelligence (LeCun et al., 2006), and biological versions of deep learning (Scellier & Bengio, 2017; Richards et al., 2019; Song et al., 2022). As a continuous-time theory of cortical processing that instantaneously corrects neuronal activities throughout the network, it makes a link to optimal feedback control (Todorov & Jordan, 2002) that has recently been considered in terms of energy-based models at equilibria (Meulemans, Zucchet, et al., 2022). It can further be seen in the tradition of predictive coding (Rao & Ballard, 1999; Bastos et al., 2012), where cortical feedback connections try to explain away lower-level activities. Yet, our prospective coding extrapolates from current quantities to predict activity in the future, not the activity at the current point in time, as in classical predictive coding. The NLA principle combines energy-based models with prospective coding in which neuronal integration delays are compensated on the fly (as also done in Haider et al., 2021). The prospective coding within each neuron may overcome putative delays that extend across the whole cortex and would allow for a real-time processing of stimuli while instantaneously correcting for ongoing errors. In this sense, the NLA represents an alternative to the recently suggested forward-forward algorithm which circumvents the delay problem by avoiding error propagation altogether (Hinton, 2022).

The paper is organized as follows: we first define the prospective somato-dendritic mismatch error, construct out of this the mismatch energy of a network, and ‘minimise’ this energy to obtain the error-corrected, prospective voltage dynamics of the network neurons. We then show that the prospective error coding leads to an instantaneous and joint processing of low-pass filtered input signals and backpropagated errors. Applied to motor control, the instantaneous processing is interpreted as a moving equilibrium hypothesis according to which sensory inputs, network state, motor commands and muscle feedback are in a self-consistent equilibrium at any point of the movement. We then derive a local learning rule that globally minimizes the somato-dendritic mismatch errors across the network, and show how this learning can be implemented through error-extracting cortical microcircuits and dendritic predictive plasticity.

Results

Somato-dendritic mismatch errors and the Lagrangian of cortical circuits

We consider a network of neurons – identified as pyramidal cells – with firing rates ri(t) in continuous time t. The somatic voltage ui of pyramidal neuron i is driven by the close-by basal input current, ∑j Wijrj, with presynaptic rates rj and synaptic weights Wij, and an additional distal apical input ei that will be learned to represent a prospective prediction error at any moment in time (Fig. 1a). While in classical rate-based neuron models the firing rate ri of a neuron is a function of the somatic voltage, ρ(ui), we postulate that the firing rate is prospective and extrapolates from ρ(ui) into the future with the temporal derivative, , with representing the temporal derivative of ρ(ui(t)). There is experimental evidence for such prospective coding in cortical pyramidal neurons to which we return later (Fig. 2a).

Somato-dendritic mismatch energies and the neuronal least-action (NLA) principle.

(a1) Sketch of a cross-cortical network of pyramidal neurons described by NLA. (a2) Correspondence between elements of NLA and biological observables such as membrane voltages and synaptic weights. (b1) The NLA principle postulates that small variations δũ (dashed) of the trajectories ũ (solid) leave the action invariant, δA = 0. It is formulated in the look-ahead coordinates ũ (symbolized by the spyglass) in which ‘hills’ of the Lagrangian (shaded grey zones) are foreseen by the prospective voltage so that the trajectory can turn by early enough to surround them. (b2) In the absence of output nudging (β = 0), the trajectory u(t) is solely driven by the sensory input, and prediction errors and energies vanish (L = 0, outer blue trajectory at bottom). When nudging the output neurons towards a target voltage (β > 0), somato-dendritic prediction errors appear, the energy increases (red dashed arrows symbolising the growing ‘volcano’) and the trajectory u(t) moves out of the L = 0 hyperplane, riding on top of the ‘volcano’ (red trajectory). Synaptic plasticity reduces the somato-dendritic mismatch along the trajectory by optimally ‘shoveling down the volcano’ (blue dashed arrows) while the trajectory settles in a new place on the L = 0 hyperplane (inner blue trajectory at bottom).

Prospective coding in cortical pyramidal neurons enables instantaneous voltage-to-voltage transfer.

(a1) The instantaneous spike rate of cortical pyramidal neurons (top) in response to sinusoidally modulated noisy input current (bottom) is phase-advanced with respect to the input (adapted from Köndgen et al., 2008). (a2) Similiarly, in NLA, the instantaneous firing rate of a model neuron , black) is phase-advanced with respect to the underlying voltage (u, red, postulating that the low-pass filtered rate is a function of the voltage, ). (b) Dendritic input in the apical tree (here called ē) is instantaneously causing a somatic voltage modulation (u, (modeling data from Ulrich, 2002)). The low-pass filtering with τ along the dendritic shaft is compensated by a lookahead mechanism in the dendrite . In Ulrich (2002) a phase advance is observed even with respect to the dendritic input current, not only the dendritic voltage, although only for slow modulations (as here). (c) While the voltage of the first neuron (u1) integrates the input rates rin from the past (bottom black upward arrows), the output rate r1 of that first neuron looks ahead in time, (red dashed arrows pointing into the future). The voltage of the second neuron (u2) integrates the prospective rates r1 (top black upwards arrows). By doing so, it inverts the lookahead operation, resulting in an instantaneous transfer from u1(t) to u2(t) (blue arrow and circles).

The second central notion of the theory is the prospective somato-dendritic mismatch error in the individual network neurons, ei(t). It is defined as a mismatch between the prospective voltage, , and the weighted prospective input rates, . If the prospective error is low-pass filtered with time constant τ, it takes the form , where is the corresponding low-pass filtered firing rate of the presynaptic neuron j (Methods). We refer to ēi as somato-dendritic mismatch error of neuron i.

We next interpret the mismatch error ēi in terms of the morphology of pyramidal neurons with basal and apical dendrites. While the error is formed in the apical dendrite, this error is added to the somatic voltage and, from the perspective of the basal inputs, it becomes a somato-dendritic mismatch error. The mismatch error tells the difference between ‘what a neuron does’, which is based on the somatic voltage ui, and ‘what the basal inputs think it should do’, which is based on its own input (Fig. 1a2). The two quantities may deviate because the neuron gets the apical input ēi that integrates in the soma, i, but not in the basal dendrite. What cannot be predicted from the somatic voltage ui by the basal input remains as ‘somato-basal’ mismatch error, , and this is just the apical error ēi.

Associated with this mismatch error is the somato-dendritic mismatch energy defined for each network neuron i ∈ 𝒩 as the squared mismatch error, On a subset of output neurons of the whole network, 𝒪 ⊆ 𝒩, a cost is defined as a function of the somatic voltage and some instructive reference signal such as a targets or a reward. When a target trajectory is available, the cost is defined at each time point as a squared target error Much more general cost functions and mismatch energies are conceivable, encompassing e.g. conductance-based neurons or including further dynamic variables (see SI). The cost represents a performance measure for the entire network that produces the output voltages uo(t) in response to some input rates rin(t). The cost directly relates to behavioral or cognitive measures such as the ability of an animal or human to perform a particular task in real time. The target could be provided by explicit external supervision, for example target movements in time encoded by , it could represent an expected reward signal, or it could arise via self-supervision from other internal prediction errors.

We define the Lagrangian (or total ‘energy’) of the network as a sum across all mismatch energies and costs, weighted by the nudging strength β of the output neurons, The low-pass filtered presynaptic rates, , also encompass the external input neurons. Due to the prospective coding, the Lagrangian can be minimal at any moment in time while the network dynamics evolves. This is different from the classical predictive coding (Rao & Ballard, 1999) and energy-based approaches (Scellier & Bengio, 2017; Song et al., 2022), where a stimulus needs to be fixed in time while the network relaxes to a steady state, and only there the prediction error is minimized. Since we postulated that the firing rates are prospective and linearly extrapolate into the future, , and because any quantity can be reconstructed from its low-pass filtering at any point in time, , we conclude that the low-pass filtered rate is a function of the instantaneous somatic voltage, , see SI. As a consequence, also the Lagrangian becomes a function of the instantaneous voltage u(t) and does not depend on past voltage values, although it depends on past firing rates.

The previous mathematical operation of extrapolating from the voltage to future rates, and going back to the current voltage via low-pass filtering, is the key ingredient for the magically sounding instantaneous processing, see Fig. 2c.

To get the instantaneity it is not really necessary to precisely look into the future. It is only necessary to undo the temporal operation of the postsynaptic neuron, and this is achieved by encoding the presynaptic voltage in the prospective rate. We come back to this instantaneity, including its practical limitations, in the overnext section.

The least-action principle expressed for prospective voltages

Motivated by the prospective firing in pyramidal neurons, we postulate that cortical networks strive to look into the future to prevent instantaneous errors. Each neuron tries to move along a trajectory that minimizes its own mismatch error ēi across time (Fig. 1b1). The ‘neuronal currency’ with which each neuron ‘trades’ with others to choose its own error-minimizing trajectory is the future discounted membrane potential, These prospective voltages ũ are the ‘canonical coordinates’ entering the NLA principle, and in these prospective coordinates the overall network searches for a ‘least action trajectory’. Since from ũ we can recover the instantaneous voltage via (see SI), we can replacIe u inlthe Lagrangian and obtain L as a function of our new canonical coordinates ũ and the ‘velocities’ , i.e. , where bold fonts represent vectors. Inspired by the least-action principle from physics, we can now define the neuronal action A as a time integral of the Lagrangian, The NLA principle postulates that the trajectory ũ(t) keeps the action A stationary with respect to small variations δũ (Fig. 1b1). In other words, nature chooses a trajectory such that, when deviating a little bit from it, say by δũ, the value of A will not change (or at most up to second order in the variation), formally δA = 0. The equations of motion that keep the action stationary are known to satisfy the Euler-Lagrange equations Applying these equations to our Lagrangian yields a prospective version of the classical leaky integrator voltage dynamics, with rates r and errors e that are looking into the future (Methods), The ‘ ·’ denotes the component-wise product, and the weight matrix splits into weights from input neurons and weights from network neurons, W = (Win, Wnet). While for output neurons a target error can be defined, , for non-output neurons i no target exist and we hence set . In a control theoretic framework, the neuronal dynamics (Eq. 7a) represents the state trajectory, and the adjoint error dynamics (Eq. 7b) represents the integrated costate trajectory (Todorov, 2006).

Formally, the voltage dynamics in Eq. 7a specifies an implicit differential equation since also appears on the right-hand side. This is because the prospective rates imply through . Likewise, the prospective errors , with ē given in Eq. 7b and plugged into Eq. 7a, imply through . Nevertheless, the voltage dynamics can be run by replacing on the right-hand side of Eq. 7a by the temporal derivative from the previous time step (technically, the Hessian (1 − Wρ′ − ē′) is required to be strictly positive definite, see Methods and SI). This ensures that the voltage dynamics of Eq. 7 can be implemented in cortical neurons with a prospective firing and a prospective dendritic error (see Fig. 2).

The error expression in Eq. 7b is reminiscent of error backpropagation (Rumelhart et al., 1986) and can in fact be mathematically linked to this (Methods). Formally, the errors are backpropagated via transposed network matrix, , modulated by , the derivative of with respect to the underlying voltage. While the transpose can be constructed with various local methods (see Akrout et al., 2019; Max et al., 2022) in our simulations we mainly adhere to the phenomenon of feedback alignment (Lillicrap, Cownden, et al., 2016) and consider fixed and randomized feedback weights B (unless stated differently). Recent control theoretical work is exploiting the same prospective coding technique as expressed in Eq. 7 to tackle general time-varying optimization problems (see Simonetto et al., 2020 for a review and the SI for the detailed connection).

Prospective coding in neurons and instantaneous propagation

The prospective rates and errors entering via r and e in the NLA (Eq. 7) are consistent with the prospective coding observed in cortical pyramidal neurons in vitro (Köndgen et al., 2008). Upon sinusoidal current injection into the soma, the somatic firing rate is advanced with respect to its voltage (Fig. 2a), effectively compensating for the delay caused by the current integration. Likewise, sinusoidal current injection in the apical tree causes a lag-less voltage response in the soma (Fig. 2b, Ulrich, 2002). While the rates and errors in general can be reconstructed from their low-pass filterings via and , they become prospective in time because and ē are themselves instantaneous functions of the voltage u, and hence r and e depend on . The derivative of the membrane potential implicitly also appears in the firing mechanism of Hodgkin-Huxley-type conductances, with a quick depolarization leading to a stronger sodium influx due to the dynamics of the gating variables (Hodgkin & Huxley, 1952). This advances the action potential as compared to a firing that would only depend on u, not , giving an intuition of how such a prospective coding may arise. A similar prospective coding has been observed for retinal ganglion cells (Palmer et al., 2015) and cerebellar Purkinje cells (Ostojic et al., 2015), making a link from the visual input to the motor control.

To understand the instantaneous propagation through the network, we low-pass filter the dynamic equation (obtained by rearranging Eq. 7a), with ē given by Eq. 7b, to obtain the somatic voltage . At any point in time, the voltage is in a moving equilibrium between forward and backpropagating inputs. Independently of the network architecture, whether recurrent or not, the output is an instantaneous function of the low-pass filtered input and a putative correction towards the target, , see Fig. 2c and Methods. The mapping again expresses an instantaneous propagation of voltages throughout the network in response to both, the low-pass filtered input and feedback error . This instantaneity is independent of the network size, and in a feed-forward network is independent of its depths (see also Haider et al., 2021, where the instantaneity is on the rates, not the voltages). In the absence of the look-ahead activity, each additional layer slows down the network relaxation time.

Notice that an algorithmic implementation of the time-continuous dynamics of a N -layer feedforward network would still need N calculation steps until information from layer 1 reaches layer N. However, this does not imply that an analog implementation of the prospective dynamics will encounter delays. To see why, consider a finite step-change Δu1 in the voltage of layer 1. In the absence of the look-ahead, Δu1 were mapped within the infinitesimal time interval dt to an infinitesimal change du2 in the voltages of layer 2. But with a prospective firing rate, , a step-change Δu1 translates to a delta-function in r1, this in turn to a step-change in the low-pass filtered rates , and therefore within dt to a step-change Δu2 in the voltages u2 of the postsynaptic neurons (Fig. 2c). Iterating this argument, a step-change Δu1 propagates ‘instantaneously’ through N layers within the ‘infinitesimal’ time interval N dt to a step-change ΔuN in the last layer. When run in a biophysical device in continuous time that exactly implements the dynamical equations (Eq. 7), the implementation becomes an instantaneous computation. Yet, in a biophysical device information has to be moved across space. This typically introduces further propagation delays that may not be captured in our formalism where low-pass filtering and prospective coding cancel each other exactly. Nevertheless, analog computation in continuous time, as formalized here, offers an idea to ‘instantaneously’ realize an otherwise time consuming numerical recipe run on time-discrete computing systems that operate with a finite clock cycle.

Prospective control and the moving equilibrium hypothesis

Crucially, at the level of the voltage dynamics (Eq. 7a) the correction is based on the prospective error e. This links our framework to optimal control theory and motor control where delays are also taken into account, so that a movement can be corrected early enough (Wolpert & Ghahramani, 2000; Todorov & Jordan, 2002; Todorov, 2004). The link between energy-based models and optimal control was recently drawn for strong nudging (β → ∞) to learn individual equilibrium states (Meulemans, Zucchet, et al., 2022). Our prospective error e(t) appears as a ‘controller’ that, when looking at the output neurons, pushes the voltage trajectories towards the target trajectories. Depending on the nudging strength β, the control is tighter or weaker. For infinitely large β, the voltages of the output neurons are clamped to the time-dependent target voltages, (implying ), while their errors, , instantaneously correct all network neurons. For small β, the output voltages are only weakly controlled, and they are dominated by the forward input, .

In the context of motor control, our network mapping can be seen as a forward internal model that quickly calculates an estimate of the future muscle length uo based on some motor plans, sensory inputs, and the current proprioceptive feedback (Fig. 3a). Forward models help to overcome delays in the execution of the motor plan by predicting the outcome, so that the intended motor plans and commands can be corrected on the fly (Kawato, 1999; Wolpert & Ghahramani, 2000).

Moving equilibrium hypothesis for motor control and real-time learning of cortical activity.

(a) A voluntary movement trajectory can be specified by the target length of the muscles in time, , encoded through the γ-innervation of muscle spindles, and the deviation of the effective muscle lengths from the target, . The Ia-afferents emerging from the spindles prospectively encode the error, so that their low-pass filtering is roughly proportional to the length deviation, truncated at zero (red). The moving equilibrium hypothesis states that the low-pass filtered input , composed of the movement plan and the sensory input (here encoding the state of the plant e.g. through visual and proprioceptive input, and ), together with the low-pass filtered error feedback from the spindles, , instantaneously generate the muscle lengths, , and are thus at any point in time in an instantaneous equilibrium (defined by Eq. 7). (b1) Intracortical iEEG activity recorded from 56 deep electrodes and projected to the brain surface. Red nodes symbolize the 56 iEEG recording sites modeled alternately as input or output neurons, and blue nodes symbolize the 40 ‘hidden’ neurons for which no data is available, but used to reproduce the iEEG activity. (b2) Corresponding NLA network. During training, the voltages of the output neurons were nudged by the iEEG targets (black input arrows, but for all red output neurons). During testing, nudging was removed for 14 out of these 56 neurons (here, represented by neurons 1, 2, 3). (c1) Voltage traces for the 3 example neurons in a2, before (blue) and after (red) training, overlaid with their iEEG target traces (grey). (c2) Total cost, integrated over a window of 8 s of the 56 output nodes during training with sequences of the same duration. The cost for the test sequences was evaluated on a 8 s window not used during training.

The observation that muscle spindles prospectively encode the muscle length and velocity (Dimitriou & Edin, 2010)) suggests that the prospective coding in the internal forward model mirrors the prospective coding in the effective forward pathway. This forward pathway leads from motor plan to spindle feedback, integrating also cerebellar and brainstem feedback (Kawato, 1999). Based on the motor plans, the intended spindle lengths and the effective muscle innervation are communicated via descending pathway to activate the γ-and α-motoneurons, respectively (Li et al., 2015). The transform from the intended arm trajectory to the intended spindle lengths via γ-innervation is mainly determined by the joint geometry. The transform from the intended arm trajectory to the force generating α-innervation, however, needs to also take account of the internal and external forces, and this is engaging our network W.

When we prepare an arm movement, spindles in antagonistic muscles pairs that measure the muscle length are tightened or relaxed before the movement starts (Papaioannou & Dimitriou, 2021). According to the classical equilibrium-point hypothesis (Feldman & Levin, 2009; Latash, 2010), top-down input adjusts the activation threshold of the spindles through (γ-)innervation from the spinal cord so that slight deviations from the equilibrium position can be signaled (Fig. 3a). We postulate that this γ-innervation acts also during the movement, setting an instantaneous target for the spindle lengths. The effective lengths of the muscle spindles is uo, and the spindles are prospectively signaling back the deviation from the target through the Ia-afferents (Dimitriou & Edin, 2010; Dimitriou, 2022). The low-pass filtered Ia-afferents may be approximated by a threshold-nonlinearity, , with β being interpreted as spindle gain (Latash, 2018). Combining the feedback from agonistic and antagonistic muscle pairs allows for extracting the scaled target error . Taking account of the prospective feedback, we postulate the moving equilibrium hypothesis according to which the instructional inputs, , the spindle feedback, , and the muscle lengths, uo, are at any point of the movement in a dynamic equilibrium. The moving equilibrium hypothesis extends the classical equilibrium-point hypothesis from the spatial to the temporal domain.

With regard to the interpretation of the prospective feedback error as spindle activity, it is worth noticing that in humans the spindle activity is not only ahead of the muscle activation (Dimitriou & Edin, 2010), but also shares the property of a motor error (Dimitriou, 2016). The experiments show that during the learning of a gated hand movement, spindle activity is initially stronger when making movement errors, and it returns back to baseline with the success of learning. We next address how the synaptic strengths W specifying the forward model can be optimally adapted to capture this learning.

Local plasticity at basal synapses minimizes the global cost in real time

The general learning paradigm starts with input time series rin(t),i and target time series, while assuming that the target series are an instantaneous function of the low-pass filtered input series, . The low-pass filtering in the individual inputs could be with respect to any time constant (that may also be learned, see SI), but for the network neurons we assume the same time constants for the voltage integration and the prospective rate. The goal of learning is to adapt the synaptic strengths W in the student network so that this approximates the target mapping, FW ≈ F*. This will also reduce the cost C defined on the output neurons in terms of the deviation of the voltage from the target, (Eq. 2).

The problem of changing synaptic weights to correct the behavior of downstream neurons, potentially multiple synapses away, is typically referred to as the credit assignment problem and is notoriously challenging in physical or biological substrates operating in continuous time. A core aspect of the NLA principle is how it relates the cost C to the Lagrangian L and eventually to somato-dendritic prediction errors ē that can be reduced through synaptic plasticity . We define this synaptic plasticity as partial derivative of the Lagrangian with respect to the weights, . Since the somato-dendritic mismatch error is , this leads to the local learning rule of the form ‘postsynaptic error times low-pass filtered presynaptic rate’, This plasticity rule runs simultaneously to the neuronal dynamics in the presence of a given nudging strength β that tells how strongly the voltage of an output neurons is pushed towards the target, . The learning rule is local in space since is represented as voltage of the basal dendrites, and the somatic voltage u may be read out at the synaptic site on the basal dendrite from the backpropagating action potentials that sample u at a given time (Urbanczik & Senn, 2014). The basal voltage becomes the dendritic prediction of the somatic activity u, interpreting Eq. 8 as ‘dendritic predictive plasticity’.

We have derived the neuronal dynamics as a path that keeps the action stationary. Without external teaching signal, the voltage trajectory wriggles on the bottom of the energy landscape (L = 0, Fig. 1b1). If the external nudging is turned on, β > 0, errors emerge and hills grow out of the landscape. The trajectory still tries to locally minimize the action, but it is lifted upwards on the hills (L > 0, Fig. 1b2). Synaptic plasticity reshapes the landscape so that, while keeping β fixed, the errors are reduced and the landscape again flattens. The transformed trajectory settles anew in another place (inside the ‘volcano’ in Fig. 1b2). Formally, the local plasticity rule (Eq. 8) is shown to perform gradient descent on the Lagrangian. In the energy landscape picture, plasticity ‘shovels off’ earth along the voltage path so that this is lowered most efficiently. The error that is back-propagated through the network tells at any point on the voltage trajectory how much to ‘dig’ in each direction, i.e. how to adapt the basal input in each neuron in order to optimally lower the local error.

The following theorem tells that synaptic plasticity pushes the network mapping towards the target mapping at any moment in time. The convergence of the mapping is a consequence of the fact the plasticity reduces the Lagrangian L = EM + βC along its gradient.

(real-time Dendritic Error Propagation, rt-DeEP)

Consider an arbitrary network W with voltage and error dynamics following Eq. 7. Then the local plasticity rule (Eq. 8), acting at each moment along the voltage trajectories, is gradient descent

on the Lagrangian L for any nudging strength β ≥ 0, i.e. , with .

on the cost C for small nudging, β → 0, while up-scaling the error to , i.e. .

The gradient statements hold at any point in time (long enough after initialization), even if the input trajectories rin(t) contain delta-functions and the target trajectories contain step-functions.

Loosely speaking, the NLA enables the network to localize an otherwise global problem: what is good for a single neuron becomes good for the entire network. In the limit of strong nudging (β→ ∞), the learning rule performs gradient descent on the mismatch energies in the individual neurons in the network. If the network architecture is powerful enough so that after learning all the mismatch energies vanish, , then the cost will also vanish, . This is because for the output neurons the mismatch error includes the target error. In the limit of weak nudging (β →0), the learning rule performs gradient descent on C, and with this also finds a local minimum of the mismatch energies.

In the case of weak nudging and a single equilibrium state, the NLA algorithm reduces to the Equilibrium Propagation algorithm (Scellier & Bengio, 2017) that minimizes the cost C for a constant input and a constant target. In the case of strong nudging and a single equilibrium state, the NLA principle reduces to the Least Control Principle (Meulemans, Zucchet, et al., 2022) that minimizes the mismatch energy EM for a constant input and a constant target. While in the Least Control Principle the inputs and outputs are clamped to fixed values, the output errors are backpropagated and the network equilibrates in a steady state where the corrected network activities reproduce the clamped output activities. This state is called the ‘prospective configuration’ in (Song et al., 2022) because neurons deep in the network are informed about the distal target and are correspondingly adapted to be consistent with this distal target. In the NLA principle, after an initial transient, this relaxation process occurs on the fly while inputs and targets dynamically change, and hence the network moves along a continuous sequence of prospective configurations.

In the motor control example, the theorem tells that a given target motor trajectory is learned to be re-produced by the forward model through the dendritic predictive plasticity (Eq. 8), applied in the various layers of the forward model. We give an example below for such a learning that involves different sensory modalities to train the forward model.

Reproducing intracortical EEG recordings and recognizing handwritten digits

To illustrate the framework, we consider a recurrently connected network that learns to represent intracortical electroencephalogram (iEEG) data from epileptic patients (Fig. 3b). For each electrode, we assign a neuron within this network to represent the activity of the cell cluster recorded in the corresponding iEEG signal via its membrane potential. During learning, a randomly selected subset of electrode neurons are nudged towards the target activity from recorded data while learned to be reproduced by the other neurons. After learning, we can present only a subset of electrode neurons with previously unseen recordings and observe how the activity of the other neurons closely matches the recordings of their respective electrodes (Fig. 3c). The network derived from NLA is thus able to learn complex correlations between signals evolving in real time by encoding them into its connectivity structure.

As an example of visual processing in the NLA framework, we next consider a well-studied image recognition task, here reformulated in a challenging time-continuous setting, and interpreted as motor task where 1 out of 10 fingers has to be bent upon seeing a corresponding visual stimulus (see Fig. 3a). In the context of our moving equilibrium hypothesis, we postulate that during the learning phase, but not the testing phase, an auditory signal identifying the correct finger sets the target spindle lengths of the 10 finger flexors, , i.e. a desired contraction for the correct finger in response to the visual input rin(t), and a desired relaxation for the 9 incorrect fingers.

We train a hierarchical three-layer network on images of handwritten digits (MNIST, LeCun, 1998), with image presentation times between 0.5τ (= 5 ms, with τ the membrane time constant) and 20τ (= 200 ms). Fig. 4a-c1 depict the most challenging scenario with the shortest presentation time. Synaptic plasticity is continuously active, despite the network never reaching a steady state (Fig. 4b1). Due to the lookahead firing rates in NLA, the mismatch errors ēi(t) propagate without lag throughout the network, as explained above. As a consequence, our mismatch errors are equal to the errors obtained from classical error backpropagation applied at each time step to the purely forward network, without error-correcting the voltage, and instead, at each layer l considering only the forward voltage updates ul = Wl ρ(ul−1), see Fig. 4b2. The network eventually learned to implement the mapping with a performance comparable to error-backpropagation at each dt, despite the short presentation time of only 5 ms (Fig. 4c1). The limiting stimulus presentation time in the order of τin (Fig. 4c2) is own to the fact that for simultaneous step-changes in the input rates and the target we required , instead of . The time constant for low-pass filtering the input rates was chosen to be τin = 10 ms, like the time constants τ for the prospective rates and the downstream membranes, but τin could have also been chosen much smaller to improve performance at even shorter presentation times.

On-the-fly learning of finger responses to visual input with real-time Dendritic Error Propagation (rt-DeEP).

(a) Functionally feedforward network with handwritten digits as visual input ( in Fig. 3a, here from the MNIST data set, 5 ms presentation time per image), backprojections enabling credit assignment, and activity of the 10 output neurons interpreted as commands for the 10 fingers (forward architecture: 784× 500 ×10 neurons). (b) Example voltage trace (b1) and local error (b2) of a hidden neuron in NLA (red) compared to an equivalent network without lookahead rates (orange). Note that neither network achieves a steady state due to the extremely short input presentation times. Errors calculated via exact backpropagation, i.e. by using the error backpropagation algorithm on a purely feedforward NLA network at every simulation time step (with output errors scaled by β), shown for comparison (blue). (c) Comparison of network models during and after learning. Color scheme as in (b). (c1) The test error under NLA evolves during learning on par with classical error backpropagation performed each Euler dt based on the feedforward activities. In contrast, networks without lookahead rates are incapable of learning such rapidly changing stimuli. (c2) With increasing presentation time, the performance under NLA further improves, while networks without lookahead rates stagnate at high error rates. This is caused by transient, but long-lasting misrepresentation of errors following stimulus switches: when plasticity is turned off during transients and is only active in the steady state, comparably good performance can be achieved (dashed orange). (d) Receptive fields of 6 hidden-layer neurons after training, demonstrating that even for very brief image presentation times (5ms), the combined neuronal and synaptic dynamics are capable of learning useful feature extractors such as edge filters.

The instantaneous voltage propagation reduces an essential constraint of previous models of bio-plausible error backpropagation (e.g., Scellier & Bengio (2017), Whittington & Bogacz (2017), and Sacramento et al. (2018), with reviews Richards et al. (2019), Whittington & Bogacz (2019), and Lillicrap, Santoro, et al. (2020)): without lookahead firing rates, networks need much longer to correctly propagate errors across layers, with each layer roughly adding another membrane time constant of 10 ms, and thus cannot cope with realistic input presentation times. In fact, in networks without lookahead output, learning is only successful if plasticity is switched off while the network dynamics is not stationary (Fig. 4c2). As a comparison, neuronal response latency in the primary visual cortex (V1) of rats to flashing stimuli are in the order of 50 ms if the cortex is in a synchronized state, shortens to roughly 40 ms if in a desynchronized state (Wang et al., 2014), and potentially shortens further if the area is preactivated through expectations (Blom et al., 2020).

Implementation in cortical microcircuits

So far, we did not specify how errors e appearing in the differential equation for the voltage (Eq. 7a) are transmitted across the network in a biologically plausible manner. Building on Sacramento et al., 2018, we propose a cortical microcircuit to enable this error transport, with all neuron dynamics evolving according to the NLA principle. Although the idea applies for arbitrarily connected networks, we use the simpler case of functionally feedforward networks to illustrate the flow of information in these microcircuits (Fig. 5a).

Hierarchical plastic microcircuits implement real-time Dendritic Error Learning (rt-DeEL).

(a) Microcircuit with ‘top-down’ input (originating from peripheral motor activity, blue line) that is explained away by the lateral input via interneurons (dark red), with the remaining activity representing the error ēl. Plastic connections are denoted with a small red arrow and nudging with a dashed line. (b1) Simulated network with 784-300-10 pyramidal-neurons and a population of 40 interneurons in the hidden layer used for the MNIST learning task where the handwritten digits have to be associated to the 10 fingers. (b2) Test errors for rt-DeEL with joint tabula rasa learning of the forward and lateral weights of the microcircuit. A similar performance is reached as with classical error backpropagation. For comparability, we also show the performance of a shallow network (dashed line). (b3) Angle derived from the Frobenius norm between the lateral pathway and the feedback pathway BlWl+1. During training, both pathways align to allow correct credit assignment throughout the network. Indices are dropped in the axis label for readability.

For such an architecture, pyramidal neurons in area l (that is a ‘layer’ of the feedforward network) are accompanied by a pool of interneurons in the same layer (area). The dendrites of the interneurons integrate in time (with time constant τ) lateral input from pyramidal neurons of the same layer (rl) through plastic weights . Additionally, interneurons receive ‘top-down nudging’ from pyramidal neurons in the next layer through randomly initialized and fixed backprojecting synapses targeting the somatic region, and interneuron nudging strength βI. The notion of ‘top-down’ originates from the functionally feed-forward architecture leading from sensory to ‘higher cortical areas’. In the context of motor control, the highest ‘area’ is last stage controlling the muscle lengths, being at the same time the first stage for the proprioceptive input (Fig. 3a).

According to the biophysics of the interneuron, the somatic membrane potential becomes a convex combination of the two types of afferent input (Urbanczik & Senn, 2014), In the biological implementation, the feedback input is mediated by the low-pass filtered firing rates , not by ul+1 as expressed in the above equation. Yet, we argue that for a threshold-linear ρ the ‘top-down nudging’ by the rate is effectively reduced to a nudging by the voltage ul+1. This is because errors are only backpropagated when the slope of the transfer function is positive, , and hence when the upper-layer voltage is in the linear regime. For more general transfer functions, we argue that short-term synaptic depression may invert the low-pass filtered presynaptic rate back to the presynaptic membrane potential, , provided that the recovery time constant τ matches the membrane time constant (see end of the section and Methods).

Apical dendrites of pyramidal neurons in each layer receive top-down input from the pyramidal population in the upper layer through synaptic weights Bl. These top-down weights could be learned to predict the lower-layer activity (Rao & Ballard, 1999) or to become the transposed of the forward weight matrix (Max et al., 2022), but for simplicity we randomly initialized them and keep them fixed (Lillicrap, Santoro, et al., 2020). Beside the top-down projections the apical dendrites also receive lateral input via an interneuron population in the same layer through synaptic weights that are plastic and will be learned to obtain suitable dendritic errors. The ‘-’ sign is suggestive for these interneurons to subtract away the top-down input entering through Bl (while the weights can still be positive or negative). Assuming again a conversion of rates to voltages, also for the inhibitory neurons that may operate in a linear regime, the overall apical voltage becomes What cannot be explained away from the top-down input Blul+1 by the lateral feedback, , remains as dendritic prediction error in the apical tree (Fig. 5a). If the top-down and lateral feedback weights are learned as outlined next, these apical prediction errors take the role of the backpropagated errors in the classical backprop algorithm.

To adjust the interneuron circuit in each layer (‘area’), the synaptic strengths from pyramidal-to-interneurons, , are learned to minimize the interneuron mismatch energy, . The interneurons, while being driven by the lateral inputs , learn to reproduce the upper-layer activity that also nudges the interneuron voltage. Learning is accomplished if the upper-layer activity, in the absence of an additional error on the upper layer, is fully reproduced in the interneurons by the lateral input.

Once the interneurons learned to represent the ‘error-free’ upper-layer activity, they can be used to explain away the top-down activities that also project to the apical trees. The synaptic strengths from the inter-to-pyramidal neurons, , are learned to minimize the apical mismatch energy, . While in the absence of an upper-layer error, the top-down activity Blul+1 can be fully cancelled by the interneuron activity , a neuron-specific error will remain in the apical dendrites of the lower-level pyramidal neurons if there was an error endowed in the upper-layer neurons. Gradient descent learning on these two energies results in the learning rules for the P-to-I and I-to-P synapses, The following theorem on dendritic error learning tells that the plasticity in the lateral feedback loop leads to an appropriate error representation in the apical dendrites of pyramidal neurons.

(real-time Dendritic Error Learning, rt-DeEL)

Consider a cortical microcircuit composed of pyramidal and interneurons, as illustrated in Fig. 5a (with dimensionality constraints specified in Methods, and fixed or dynamic weights to the interneurons). Then the inter-to-pyramidal synapses evolve at each layer l (‘cortical area’) such that the lateral feedback circuit aligns with the top-down feedback circuit, In the presence of an output nudging, the apical voltages of the layer-l pyramidal neurons (Eq. 10) then represent the ‘B-backpropagated’ error . When modulated by the postsynaptic rate derivative, , this apical error ensures the correct representation of errors ēl for the real-time dendritic error propagation (rt-DeEP, Theorem 1),

Simultaneously learning apical errors and basal signals

Microcircuits following these neuronal and synaptic dynamics are able to learn the classification of hand-written digits from the MNIST dataset while learning the apical signal representation (Fig. 5b1-2). In this case, feedforward weights Wl and lateral weights and are all adapted simultaneously. Including the -plasticity (by turning on the interneuron nudging from the upper layer, βI > 0 in Eq. 9), greatly speeds up the learning.

With and without -plasticity, the lateral feedback via interneurons (with effective weigh ) learns to align with the forward-backward feedback via upper layer pyramidal neurons (with effective weight BlWl+1, Fig. 5b3). The microcircuit extracts the gradient-based errors (Eq. 13), while the forward weights use these errors to reduce these errors to first minimize the neuron-specific mismatch errors, and eventually the output cost.

Since the apical voltage appears as a postsynaptic factor in the plasticity rule for the interneurons , this I-to-P plasticity can be interpreted as Hebbian plasticity of inhbitory neurons, consistent with earlier suggestions (Vogels et al., 2012; Bannon et al., 2020). The plasticity of the P-to-I synapses, in the same way as the plasticity for the forward synapses s, can be interpreted as learning from the dendritic prediction of somatic activity (Urbanczik & Senn, 2014).

Crucially, choosing a large enough interneuron population, the simultaneous learning of the lateral microcircuit and the forward network can be accomplished without fine-tuning of parameters. As an instance in case, all weights shared the same learning rate. Such stability bolsters the biophysical plausibility of our NLA framework and improves over the previous, more heuristic approach (Sacramento et al., 2018; Mesnard et al., 2019). The stability may be related to the nested gradient descent learning according to which somatic and apical mismatch errors in pyramidal neurons, and somatic mismatch errors in inhibitory neurons are minimized.

Finally, since errors are defined at the level of membrane voltages (Eq. 11), synapses need a mechanism by which they can recover the presynaptic voltage errors from their afferent firing rates. While for threshold-linear transfer functions this is not too involved (Methods), more general neuronal nonlinearities must be matched by corresponding synaptic nonlinearities. Pfister et al. (2010) have previously illustrated how spiking neurons can leverage short-term synaptic depression to estimate the membrane potential of their presynaptic partners. Here, we assume a similar mechanism in the context of our rate-based neurons. The monotonically increasing neuronal activation function, , can be approximately compensated by a vesicle release probability that monotonically decreases with the low-pass filtered presynaptic rate (see SI and Fig. 6 therein). If properly matched, this leads to a linear relationship between the presynaptic membrane potential ul+1 and the postsynaptic voltage contribution.

Recovering presynaptic potentials through short term depression.

(a1) Relative voltage response of a depressing cortical synapse (recreated from Abbott et al., 1997), identified as synaptic release probability p. (a2) The product of the low-pass filtered presynaptic firing rate times the synaptic release probability is proportional to the presynaptic membrane potential, . (a3) Average in vivo firing rate of a neuron in the visual cortex as a function of the somatic membrane potential (recreated from Anderson et al., 2000), which can be qualitatively identified as the stationary rate derived in Eq. 43.

Discussion

We introduced a least-action principle for neuronal networks from which we derived the membrane potential dynamics of the involved neurons, together with gradient descent plasticity of the synapses in the network. The central notion of the theory is the somato-dendritic mismatch error in each individual neuron. Given the various findings on the dendritic integration of top-down signals (Larkum, 2013; Takahashi et al., 2020), we suggest that the error is formed in the apical dendrite of pyramidal neurons. Active dendritic and somatic processes compensate for delays caused by the dendritic error propagation and the somatic integration. This leads to the prospective firing of pyramidal neurons that are jointly driven by the forward input on the basal dendrites and the error on the apical dendrite. We derived a local plasticity rule for synapses on the basal tree that, by extracting the apical feedback error, reduces the prospective somato-dendritic mismatch error in an individual neuron. This plasticity is shown to also globally reduce the instantaneous cost at output neurons (defined by deviations from target voltages) at any moment in time while stimuli and targets may continuously change. Motor control offers an intuitive context where self-correcting prospective networks may enter. We put forward the moving equilibrium hypothesis, according to which sensory input, motor commands and muscle spindle feedback are in a recurrent equilibrium at any moment during the movement.

We further showed how a local microcircuit can serve to calculate a neuron-specific error in the apical tree of each individual pyramidal neuron. The apical tree of pyramidal neurons receives feedback from higher-level neurons in the network, while local inhibitory neurons try on the fly to ‘explain away’ the top-down expectations. What cannot be explained away remains as apical prediction error. This prediction error eventually induces error-correcting plasticity at the sensory-driven input on the basal tree. To find the correct error representation in the apical tree, plasticity of the inter-to-pyramidal neurons seeks to homeostatically drive the apical de-or hyperpolarized voltage back to rest. While the network synapses targetting the basal tree are performing gradient descent on the global cost, the microcircuit synapses involved in the lateral feedback are gradient descent on local error functions, both at any moment in time.

Our work builds on three general lines of research on formalizing the learning capabilities of cortical networks. The first line refers to the use of an energy function to jointly infer the neuronal dynamics and synaptic plasticity, originally formulated for discrete-time networks (Hopfield, 1982; Ackley et al., 1985), and recently extended to continuous-time networks (Scellier & Bengio, 2017). The second line refers to understanding error-backpropagation in the brain (Rumelhart et al., 1986; Xie & Seung, 2003; Whittington & Bogacz, 2017; Whittington & Bogacz, 2019; Lillicrap, Santoro, et al., 2020). The third line refers to the use of dendritic compartmentalization in various kinds of computation (Schiess et al., 2016; Poirazi & Papoutsi, 2020), recently linked to deep learning (Guerguiev et al., 2017; Sacramento et al., 2018; Haider et al., 2021).

With regard to energy functions, the NLA principle adds a variational approach to characterize continuous-time neuronal trajectories and plasticity. Variational approaches are studied in the context of optimal control theory where a cost integral is minimized across time, constrained to some network dynamics (Todorov & Jordan, 2002; Meulemans, Farinha, et al., 2021). The NLA represents a unifying notion that allows to infer both, the network dynamics and its optimal control from a single Lagrangian. The error we derive represents prospective control variables that are applied to the voltages of each network neurons so that they push the output neurons towards their target trajectory. The full expression power of this control theoretic framework has yet to be proven when it is extended to genuine temporal processing that includes longer time constants, for instance inherent in a slow threshold adaptation (Bellec et al., 2020). The NLA principle can also treat the case of strong feedback studied so far in relaxation networks only (Meulemans, Zucchet, et al., 2022; Song et al., 2022). Our rt-DeEP Theorem makes a statement for real-time gradient descent learning while the network is in a moving equilibrium, linking to motor learning in the presence of perturbing force fields (Herzfeld et al., 2014) or perturbing visual inputs (Dimitriou, 2016).

With regard to error-backpropagation, the NLA principle relies on feedback alignment (Lillicrap, Cownden, et al., 2016) to correctly interpret the apical errors. Other works have explored learning of feedback weights (Akrout et al., 2019; Kunin et al., 2020; Max et al., 2022). Since these are complementary to the principles suggested here, a promising direction would be to explore how feedback weights can be learned in NLA microcircuits to improve credit assignment in deeper networks.

With regard to dendritic and neuronal processing, we emphasize the prospective nature of apical voltages and somatic firing (Fig. 2). The prospective coding of errors and signals overcomes the various delays inherent in our previous approach (Sacramento et al., 2018; Mesnard et al., 2019). Each neuron of the network instantaneously corrects its voltage so that the output neurons are pushed towards the target trajectories. Ongoing synaptic plasticity adjusts the synaptic strengths so that the errors in each neuron are reduced at any moment in time, without needing to wait for network relaxations (Scellier & Bengio, 2017; Song et al., 2022). Yet, because in the present framework the input rates are still low-pass filtered, step inputs cannot be instantaneously propagated, only their low-pass filterings. Haider et al., 2021 offers a framework that is also suited instantaneously process step inputs.

Motivated by the predictive power of the least-action principle in physics, we ask about experimental confirmation and predictions of the NLA in biology. Given its axiomatic approach, it appears astonishing to find various preliminary matches at the dendritic, somatic, interneuron, synaptic and even behavioural level. Some of these are: (1) the prospective coding of pyramidal neuron firing (Köndgen et al., 2008); (2) the prospective processing of apical signals while propagating to the soma (Ulrich, 2002); (3) the basal synaptic plasticity on pyramidal neurons and synaptic plasticity on interneurons, driven by the postsynaptic activity that is ‘unexplained’ by the distal dendritic voltage (Urbanczik & Senn, 2014); (4) the Hebbian homeostatic plasticity of interneurons targeting the apical dendritic tree of pyramidal neurons (Bannon et al., 2020); (5) the short-term synaptic depression at top-down synapses targetting inhibitory neurons and apical dendrites (akin to Abbott et al., 1997, but with a faster recovery time constant) that invert the presynaptic activation function (see also Pfister et al., 2010); (6) the modulation of the apical contribution to the somatic voltage by the slope of the somatic activation function (for instance by downregulating apical NMDA receptors with increasing rate of backpropagating action potentials, Theis et al., 2018); and (7) the role of muscle spindles in the prospective encoding of motor errors during motor learning (Dimitriou, 2016; Papaioannou & Dimitriou, 2021). More theoretical and experimental work is required to explore these various links.

Overall, our approach introduces a method from theoretical physics to computational neuroscience and couples it with a normative perspective on the dynamical processing of neurons and synapses within global cortical networks and local microcircuits. Given its physical underpinnings, the approach may inspire the rebuilding of computational principles of cortical neurons and circuits in neuromorphic hardware (Bartolozzi et al., 2022). A step in this direction, building on the instantaneous computational capabilities with slowly integrating neurons, has been done (Haider et al., 2021) and a next challenge is to generalize the NLA principle to spiking neurons (Zenke & Ganguli, 2018; Göltz et al., 2021) and longer temporal processing.

eLife Assessment

This manuscript describes a potentially important theoretical framework to link predictive coding, error-based learning, and neuronal dynamics. The provided evidence is solid but would be made more robust if the different lines of argument were more directly connected. Improving the exposition of the manuscript would make it more accessible to a broader audience.

Reviewer #1 (Public Review):

The manuscript considers a hierarchical network of neurons, of the type that can be found in the sensory cortex, and assumes that they aim to constantly predict sensory inputs that may change in time. The paper describes the dynamics of neurons and rules of synaptic plasticity that minimize the integral of prediction errors over time.

The manuscript describes and analyses the model in great detail, and presents multiple and diverse simulations illustrating the model's functioning. However, the manuscript could be made more accessible and easier to read. The paper may help to understand the organization of cortical neurons, their properties, as well as the function of their particular components (such as apical dendrites).

Reviewer #2 (Public Review):

Neuroscientists often state that we have no theory of the brain. The example of theoretical physics is often cited, where numerous and quite complex phenomena are explained by a compact mathematical description. Lagrangian and Hamiltonian pictures provide such powerful 'single equation'. These frameworks are referred to as 'energy', an elegant way to turn numerous differential equations into a single compact relationship between observable quantities (state variables like position and speed) and scaling constants (like the gravity constant or the Planck constant). Such energy-pictures have been used in theoretical neuroscience since the 1980s.

The manuscript "neuronal least-action principle for real-time learning in cortical circuits" by Walter Senn and collaborators describes a theoretical framework to link predictive coding, error-based learning, and neuronal dynamics. The central concept is that an energy function combining self-supervised and supervised objectives is optimized by realistic neuronal dynamics and learning rules when considering the state of a neuron as a mixture of the current membrane potential and its rate of change. As compared with previous energy functions in theoretical neuroscience, this theory captures a more extensive range of observations while satisfying normative constraints. Particularly, no theory had to my knowledge related to adaptive dynamics widely observed in the brain (referred to as prospective coding in the text, but is sometimes referred to as adaptive coding or redundancy reduction) with the dynamics of learning rules.

The manuscript first exposes the theory of two previously published papers by the same group on somato-dendritic error with apical and basal dendrites. These dynamics are then related to an energy function, whose optimum recovers the dynamics. The rest of the manuscript illustrates how features of this model fit either normative or observational constraints. Learning follows a combination of self-supervised learning (learning to predict the next step) and supervised learning (learning to predict an external signal). The credit assignment problem is solved by an apical-compartment projecting a set of interneurons with learning rules whose role is to align many weight matrices to avoid having to do multiplexing. An extensive method section and supplementary material expand on mathematical proofs and makes more explicit the mathematical relationship between different frameworks.

Experts would say that much of the article agglomerates previous theoretical papers by the same authors that have been published recently either in archival servers or in conference proceedings. A number of adaptations to previous theoretical results were necessary, so the present article is not easily reduced to a compendium of previous pre-prints. However, the manuscript is by no means easy to read as there are several inconsistencies and it lacks a single thread. Also, there remains a few thorny assumptions (unobserved details of the learning rules or soma-dendrites interactions), but the theory is likely going to be regarded as an important step towards a comprehensive theory of the brain.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Recommendations For The Authors):

Major comments:

(1) It is nice that the authors compared their model to the one "without lookahead" in Figure 4, but this comparison requires more evidence in my opinion, as I explain in this comment. The model without lookahead is closely related or possibly equivalent to the standard predictive coding. In predictive coding, one can make the network follow the stimulus rapidly by reducing the time constant tau. However, as the time constant decreases, the network would become unstable both in simulations (due to limited integration time step) and physical implementation (due to noise). Therefore I wonder if the proposed model has an advantage over standard predictive coding with an optimized time constant. Hence I suggest to also add a comparison between the proposed model, and the predictive coding with parameters (such as tau) optimized independently for each model. Of course, we know that the time-constant of biological neurons is fixed, but biological neurons might have had different time constants (by changing leak conductance) and such analysis could shed light on the question of why the neurons are organized the way they are.

The comparison with a predictive network for which the neuronal time constants shrink towards 0 is in fact helpful. We added two news subsections in the SI that formally compares the NLA with other approaches, Equilibrium propagation and the Latent Equilibrium, with a version of Equilibrium Propagation also covering the standard predictive coding you describe (SI, Sect.C and D). The Subsection C concludes: “In the Equilibrium propagation we cannot simply take the limit t0 since then the dynamics either disappears (when tau remains on the left, t Du  0) or explodes (when t is moved to the right, dt/ t  ∞), leading to either too small or too big jumps.”

We have also expanded the passage on the predictive coding in the main text, comparing our instantaneous network processing (up to a remaining time constant tin) with experimental data from humans (see page 10 of the revised ms). The new paragraph ends with:

“Notice that, from a technical perspective, making the time constants of individual cortical neurons arbitrarily short leads to network instabilities and is unlikely the option chosen by the brain (see SI Sect. C, Comparison to the Equilibrium Propagation).”

A new formal definition of the moving equilibrium in the Methods (Sect. F) helps to understand this notion of being in a balanced equilibrium state during the dynamics. This formal definition directly leads to the contraction analysis in the SI, Sect. D, showing why the Latent Equilibrium is always contractive, while the current form of the NLA may show jumps at the corner of a ReLu (since a second order derivative of the transfer function enters in the error propagation).

The reviewer perhaps has additional simulations in mind that compare the robustness of the different models. However, as this paper is more about presenting a novel concept with a comprehensive theory (summing up to 45 pages), we prefer to not add more than the simulations necessary to check the statements of the theorems.

(2) I found this paper difficult to follow, because the Results sections went straight into details, and various elements of the model were introduced without explaining why they are necessary. Furthermore, the neural implementation was introduced after the model simulations. I suggest reorganizing the manuscript, to describe the model following Marr's levels of description and then presenting the results of simulations. In particular, I suggest starting the Results section by explaining what computation the network is trying to achieve (describe the setup, function L, define its integral over time, and explain that the goal is to find a model minimizing this integral). Then, I suggest presenting the algorithm the neurons need to employ to minimize this integral, i.e. their dynamics and plasticity (I wonder if r=rho(u) + tau rho(u)' is a consequence of action minimization or a necessary assumption - please clarify it). Next please explain how the algorithms could be implemented in biological neurons. Afterward please present the results of the simulation.

We are sorry to realize that we could not convey the main message clearly enough. After rewriting the paper and straightening the narrative, we hope it is simpler to understand now.

The paper does not suggest a new model to solve a task, and writing down the function to be minimized is not enough. The point of the NLA is that the time integral of our Lagrangian is minimized with respect to the prospective coordinates, i.e. the discounted future voltage. It is about the question how dynamic equations in biology are derived. Of course, we also solve these equations, prove theorems and perform simulations. But the main point that biology seems to deal with time differently than physics deals with time. Biology “thinks” in terms of future quantities, physics “thinks” in terms of current quantities. We tried to explain this better now in the Introduction, the Results (e.g. after Eq. 5) and the Methods.

(3) Understanding the paper requires background knowledge that most readers of eLife are unlikely to have, even if they are mathematically minded. For example, I am from the field of computational neuroscience, and I have never heard about Least Action principle from physics or the EulerLagrange equation. I felt lost after reading this paper, and to be able to write this review I needed to watch videos on the Euler-Lagrange equation. To help other readers, I have two suggestions: First, I feel that Eq 4-6 could be moved to the methods, because I found the concept of u~ difficult to understand, and it does not appear in the algorithm. Second, I advise to write in the Introduction, what knowledge is required to follow this paper, and point the readers to resources where they can find the required information. The authors may specify what background is required to follow the main text, and what is required to understand the methods.

We hope that after explaining the rationale better, it becomes clear that we cannot skip the equations for the prospective coordinates. Likewise, the Euler-Lagrange equations need to be presented in the abstract form, since these are the equations that are eventually transformed into the “model”. We tried to give the basic intuition for this in the main text. As we explained above, the equations asked to be skipped represent the essence of the proposal. It is about how to derive a model equations.

Moreover, we give more explanations in the Methods to understand the derivations, and we refer to the specifically sections in the SI for further details. We are aware that a full understanding of the theory requires some basic knowledge of the calculus of variation.

We are hesitating to write in the Introduction what type of knowledge is required to understand the paper. An understanding can be on various levels. Moreover, the materials that are considered to be helpful depend on the background. While for some it is a Youtube, for some Wikipedia, and for others it is a textbook where specific ingredients can be extracted. But we do cite two textbooks in the Results and more in the SI, Sect. F, when referring to the principle of least action in physics and the mathematics, including weblinks.

Minor comments

Eq.3: The Authors refer to this equation as a Lagrangian. Could you please clarify why? Is the logic to minimize the energy subject to a constraint that Cost = 0?

Thanks for asking. The cost is not really a constraint, it is globally minimized, in parallel steps. We are explaining this right after Eq. 3. “We `prospectively' minimize L locally across a voltage trajectory, so that, as a consequence, the local synaptic plasticity for W will globally reduce the cost along the trajectory (Theorem 1 below).”

We were adding two sentence that explain why this function in Eq. 3 is called a Lagrangian: “While in classical energy-based approaches L is called the total energy, we call it the `Lagrangian' because it will be integrated along real and virtual voltage trajectories as done in variational calculus (leading to the Euler-Lagrange equations, see below and SI, Sect. F)”

p.4, below Eq. 5 - Please explain the rationale behind NLA, i.e. why is it beneficial that "the trajectory u˜(t) keeps the action A stationary with respect to small variations δu˜"? I guess you wish to minimize L integrated over time, but this is not evident from the text.

Hmm, yes and no. We wish to minimize the cost, and on the way there minimize the action. Since the global minimization of C is technically difficult, one looks for stationary trajectory as defined in the cited sentence, while minimizing L with respect to W, to eventually minimize the cost.

In the text we now explain after Eq. 5:

“The motivation to search for a trajectory that keeps the action stationary is borrowed from physics. The motivation to search for a stationary trajectory by varying the near-future voltages ũ instead of u is assigned to the evolutionary pressure in biology to 'think ahead of time'. To not react too late, internal delays involved in the integration of external feedback need to be considered and eventually need to be overcome. In fact, only for the 'prospective coordinates' defined by looking ahead into the future, even when only virtually, will a real-time learning from feedback errors become possible (as expressed by our Theorems below).”

Bottom of page 8. The authors say that in the case of single equilibrium and strong nudging the model reduced to the Least Control Principle. Does it also reduce to Predictive coding for supervised learning? If so, it would be helpful to state so.

Yes, in this case the prediction error in the apical dendrite becomes the one of predictive coding. We are stating this now right at the end of the cited sentence:

“In the case of strong nudging and a single steady-state equilibrium, the NLA principle reduces to the Least-Control Principle (Meulemans et al., 2022) that minimizes the mismatch energy E^M for a constant input and a constant target, with the apical prediction error becoming the prediction error from standard predictive coding (Rao & Ballard, 1999).”

In the Discussion we also added a further point (iv) to compare the NLA principle with predictive coding. Both “improve” the sensory representation, but the NLA does in favor of an output, and the predictive coding in favor of the sensory prediction itself (see Discussion).

Whenever you refer to supplementary materials, please specify the section, so it is easier for the reader to find it.

Done. Sorry to not have done it earlier. We are now also indicate specific sections when referring to the Methods.

Reviewer #2 (Recommendations For The Authors):

There are no major issues with this article, but I have several considerations that I think would greatly improve the impact, clarity, and validity of the claims.

(1) Unifying the narrative. There are many many ideas put forward in what feels like a deluge. While I appreciate the enthusiasm, as a reader I found it hard to understand what it was that the authors thought was the main breakthrough. For instance, the abstract, results, introduction, and discussion all seem to provide different answers to that question. The abstract seems to focus on the motor error idea. The introduction seems to focus on the novel prospective+predictive setup of the energy function. The discussion lists the different perks of the theory (delay compensation, moving equilibrium, microcircuit) without referring to the prospective+predictive setup of the energy function.

Thanks much for these helpful hints. Yes, the paper became an agglomerate of many ideas, also own to the fact that we wish to show how the NLA principle can be applied to explain various phenomenology in neurosicence. We now simplified the narrative to this one point of providing a novel theoretical framework for neuroscience, and explaining why this is novel and why it “suddenly works” (the prospective minimization of the energy).

As you can see from the dominating red in the revised pdf, we did fully rewrite Abstract, Introduction and Discussion under the narrative of the NLA and prospective coding.

(2) Laying out the organization of the notation clearly. There are quite a few subtle distinctions of what is meant by the different weight matrices (omnibus matrix then input vs recurrent then layered architecture), different temporal horizon formalisms (bar, not bar, tilde), different operators (L, curly L, derivative version, integral version). These different levels are introduced on the fly, which makes it harder to grasp. The fact that there are many duplicate notations for the same quantities does not help the reader. For instance u_0 becomes equal to u_N at one point (above Eq 25). Another example is the constant flipping between integrated and 'current input' pictures. So laying out the multiple layers early, making a table or a figure for the notation, or sticking with one level would help convey the idea to a wide readership.

Thanks for the hints. We included the table you suggested, but put it to the SI as it became a full page itself. We banned the curly L abbreviating the look-ahead operator.

The “change of notation” you are alluding to is tricky, though. In a recurrent layer, the index of the output neuron is called o. In a forward network with N layer, the index of the output neurons becomes the last layer N. One has to introduce the layer index l anway for the deeper layers l < N, and we found it more consistent to explain that, while switching from the recurrent to the forward network, the voltage of the output layer becomes now u_o = u_N. There are more of these examples, like the weight matrix W splitting into a intrinsic network part W_net across which errors backpropagate, and a part conveying the input, W_in, that has to be excluded when writing the backpropagation formula for general networks. Again, in the case of the feedforward networks, the notation reduces to W_l, with index l coding for the layer. Presenting the general approach and a specific example may appear as we would duplicate notations – we haven’t found a solution here.

(3) Separate the algorithm from the implementation level. I particularly struggled with separating the ideas that belonged to the algorithm level (cost function, optimization objectives) and the biophysics. The two are interwoven in a way that does not have to be. Particularly, some of the normative elements may be implemented by other types of biophysics than the authors have in mind. It is for this reason that I think that separating more clearly what belongs to the implementation and algorithm levels would help make the ideas more widely understood. On this point, a trigger point for me was the definition of the 'prospective input rates' e_i, which comes in the second paragraph.

We are very sorry to have made you thinking that the 'prospective input rates' would be e_i. The prospective input rates are r_i. The misunderstanding likely appeared by an unclear formulation from our side that is now corrected (see first and second paragraph of the Results where we introduce r_i and e_i).

From a biophysical perspective, it is quite arbitrary to define the input to be the difference between the basal input and the somatic (prospective) potential. It sounds like it comes from some unclear normative picture at this point. But the authors seem to have in mind to use the fact that the somatic potential is the sum of apical and basal input, that's the biophysical picture.

We hope to have disentangled the normative and biophysical view in the 2nd and 3rd paragraph of the Results, respectively. We introduce the prospective error ei as abstract notion in the first paragraph, while explaining that it will be interpreted as somato-dendritic mismatch error in neuron I in the next paragraph. The second paragraph contains the biophysical details with the apical and basal morphology.

(4) Experts and non-expert would appreciate an explanation of why/how the choice of state variables matters in the NLA. The prospective coding state variables cannot be said to be the naïve guess. Why does the simple u, dot{u} not work as state variables applied on the same energy function, as would be a naïve application of the Lagrangian ideas?

We are very glad for this hint to present an intuition behind the variation of the action with respect to a prospective state, instead of the state itself. The simple L(u, dot{u}) does not work because one does not obtain the first-order voltage dynamics compatible with the biophysics. We made an effort to explain the intuition to non-experts and experts in an additional paragraph right after presenting the voltage and error dynamics (Eq. 7 on page 4).

Here is how the paragraph starts (not displaying the formulas here):

“From the point of view of theoretical physics, where the laws of motion derived from the least-action principle contain an acceleration term (as in Newton's law of motion, like … for a harmonic oscillator), one may wonder why no second-order time derivative appears in the NLA dynamics. As an intuitive example, consider driving into a bend. Looking ahead in time helps us to reduce the lateral acceleration by braking early enough, as opposed to braking only when the lateral acceleration is already present. This intuition is captured by minimizing the neuronal action A with respect to the discounted future voltages ũi instead of the instantaneous voltages ui.

Keeping up an internal equilibrium in the presence of a changing environment requires to look ahead and compensate early for the predicted perturbations.

Technically, …”

More details are given in the Methods after Eq. 20. Moreover, in the last part of the SI, Sect. F, we have made the link to the least-action principle in physics more explicitly. There we show how the voltage dynamics can be derived from the physical least-action principle by including the Rayleigh dissipation (Eq. 92 and 95).

(5) Specify that the learning rules have not been observed. Though the learning rules are Hebbian, the details of the rules have not to my knowledge been observed. Would be worth mentioning as this is a sticking point of most related theories.

We agree, and we do now explicitly write in the Discussion that the learning rule still awaits to be experimentally tested.

1. Some relevant literature. Chalk et al. PNAS (2018) have explored the relationship between temporal predictive coding and Rao & Ballard predictive coding based on the parameters of the cost function. Harkin et al. eLife (2023) have shown that 'prospective coding' also takes place in the serotonergic system, while Kim ... Ma (2021) have put forward similar ideas for dopamine, both may participate in setting the cost function. Instantaneous voltage propagation is also a focus of Greedy et al. (2023). The authors cite Zenke et al. for spiking error propagation, but there are biological references to that end.

Thanks much for these hints. We do now cite the book of Gerstner & Kistler on spiking neurons, and more specifically the spike-based approach for learning to represent signals (Brendel, .., Machens, Denève, PLoS CB, 2020). Otherwise, we had difficulties to incorporate the other literature that seems to us not directly related to our approach, even when related notions come up (like predictive coding and temporal processing in Chalk et al. (2018), where various temporal coding schemes coding efficiency is studied as a function of the signal-to-noise ratio), or the apical activities in Greedy et al. (2022), where bursting, multiplexing and synaptic facilitation arises). We found it would confuse more than it would help if we would cite these papers too (we do already cite 95 papers).

(7) In the main text, theorem two is presented as proof without assumptions on the level of nudging, but the actual proof uses strong assumptions in that respect, relying on numerical ad hoc observations for the general case.

Thanks for pointing this out. We agree it is a better style to state all the critical assumptions in Theorem itself, rather than deferring them to the Methods. We now state: “Then, for suitable top-down nudging, learning rates, and initial conditions, the ….weights …evolve such that…”.

(8) In the discussion regarding error-backpropagation, it seems to me that it could be clarified that the current algorithm asks for a weight alignment between FF and FB matrices as well as between FB and interneuron circuit matrices. Whether all of these matrices can be learned together remains to be shown; neither Akrout, Kunin nor Max et al. have shown this explicitly. Particularly when there are other inputs to the apical dendrites from other areas.

Yes, it is difficult to learn to align all in parallel. Nevertheless, our simulations in fact do align the lateral and vertical circuits, at is also claimed in Theorem 2. Yet, as specified in the theorem, “for suitable learning rates” (that were all the same, but were commonly reduced after some training time, as previously explained in the Methods, Details for Fig. 5).

In the Discussion we now emphasis that, in general, simulating all the circuitries jointly from scratch in a single phase is tricky. We write:

“A fundamental difficulty arises when the neuronal implementation of the Euler-Lagrange equations requires an additional microcircuit with its own dynamics. This is the case for the suggested microcircuit extracting the local errors. Formally, the representation of the apical feedback errors first needs to be learned before the errors can teach the feedforward synapses on the basal dendrites. We showed that this error learning can itself be formulated as minimizing an apical mismatch energy. What the lateral feedback through interneurons cannot explain away from the top-down feedback remains as apical prediction error.

Ideally, while the network synapses targetting the basal tree are performing gradient descent on the global cost, the microcircuit synapses involved in the lateral feedback are performing gradient descent on local error functions, both at any moment in time.

The simulations show that this intertwined system can in fact learn simultaneously with a common learning rate that is properly tuned. The cortical model network of inter- and pyramidal neurons learned to classify handwritten digits on the fly, with 10 digit samples presented per second. Yet, the overall learning is more robust if the error learning in the apical dendrites operates in phases without output teaching but with corresponding sensory activity, as may arise during sleep (see e.g. Deperrois et al., 2022 and 2023).”

(9) The short-term depression model is assuming a slow type of short-term depression, not the fast types that are the focus of much recent experimental literature (like Campagnola et al. Science 2022).

This assumption should be specified.

Thanks for hinting to this literature that we were not aware of. We are now citing the releaseindependent plasticity (Campagnola et al. 2022) in the context of our synaptic depression model.

(10) There seems to be a small notation issue: Eq 21 combines vectors of the size of the full network (bar{e}) and the size of the readout network (bar{e}star).

Well, for notational convenience we set the target error to e*=0 for non-output neurons. This way we can write the total error for an arbitrary network neuron as the sum of the backpropagated error plus the putative target error (if the neuron is an output neuron). Otherwise we would always have to distinguish between network neuron that may be output neurons, and those that are not. We did say this in the main text, but are repeating it now again right after Eq. 21. -- Notations are often the result of a tradoff.

Version 2

Abstract

One of the most fundamental laws of physics is the principle of least action. Motivated by its predictive power, we introduce a neuronal least-action principle for cortical processing of sensory streams to produce appropriate behavioural outputs in real time. The principle postulates that the voltage dynamics of cortical pyramidal neurons prospectively minimize the local somato-dendritic mismatch error within individual neurons. For motor output neurons, it implies minimizing an instantaneous behavioural error. For deep network neurons, it implies a prospective firing to overcome integration delays and correct for possible output errors right in time. The neuron-specific errors are extracted in the apical dendrites of pyramidal neurons through a cortical microcircuit that tries to explain away the feedback from the periphery, and correct the trajectory on the fly. Any motor output is in a moving equilibrium with the sensory inputs and the motor feedback during the whole sensory-motor trajectory. Ongoing synaptic plasticity reduces the somato-dendritic mismatch error within each cortical neuron and performs gradient descent on the output cost at any moment in time. The neuronal least-action principle offers an axiomatic framework to derive local neuronal and synaptic dynamics for global real-time computation and learning in the brain and in physical substrates in general.

Introduction

Wigner’s remark about the ‘unreasonable effectiveness’ of mathematics in allowing us to understand physical phenomena (Wigner, 1959) is famously contrasted by Gelfand’s quip about its ‘unreasonable ineffectiveness’ in doing the same for biology (Borovik, 2021). Considering the component of randomness that is inherent to evolution, this may not be all that surprising. However, while this argument holds just as well for the brain at the cellular level, ultimately brains are computing devices. At the level of computation, machine learning and neuroscience have revealed near-optimal strategies for information processing and storage, and evolution is likely to have found similar principles through trial and error (Hassabis et al., 2017). Thus, we have reason to hope for the existence of fundamental principles of cortical computation that are similar to those we have found in the physical sciences. Eventually, it is important for such approaches to relate these principles back to brain phenomenology and connect function to structure and dynamics.

In physics, a fundamental measure of ‘effort’ is the action of a system, which nature seeks to ‘minimize’. Given an appropriate description of interactions between the systems constituents, the least-action principle can be used to derive the equations of motion of any physical system (Feynman et al., 2011; Coopersmith, 2017). Here, we suggest that in biological information processing, a similar principle holds for prediction errors, which are of obvious relevance for cognition and behavior. Based on such errors, we formulate a neuronal least-action (NLA) principle which can be used to derive neuronal dynamics and map them to observed dendritic morphologies and cortical microcircuits. Within this framework, local synaptic plasticity at basal and apical dendrites can be derived by stochastic gradient descent on errors. The errors that are minimized refer to the errors in output neurons that are typically thought to represent motor trajectories, planned and encoded in cortical motor areas and ultimately in the spinal cord and muscles. In the context of motor control, a phenomenological ‘minimal action principle’ has previously been proposed that guides the planning and execution of movements (Feldman & Levin, 2009). Our neuronal least-action principle reformulates and formalizes the classical equilibrium point hypothesis (Latash, 2010) in a dynamical setting, linking it to optimality principles in sensory-motor control (Todorov, 2004).

Other attempts exist to link biological information processing and neural networks with the least-action principle, for instance by directly learning to reproduce a given trajectory (Amirikian & Lukashin, 1992), by minimizing the physical action for the muscle force generation by motor unit recruitment (Senn, Wyler, Clamann, Kleinle, Larkum, et al., 1995), minimizing cognitive prediction errors (Alonso et al., 2012), minimizing output errors with a weight-change regularization (Betti & Gori, 2016), minimizing psychomotor work (Fox & Kotelba, 2018), minimizing data transport through a network (Karkar et al., 2021), minimizing the discrimination information (Summers, 2021), or minimizing the free energy (Friston, 2010; Friston et al., 2022). Apart from the latter, however, these attempts remain far from the biology that seems to resist a formalization with the tool of physics – at least, when applied too strictly.

The fundamental novelty of our NLA principle is the way it deals with time. In physics, bodies interact based on where they are now, irrespective of what happens in the future. Living systems, instead, interact based on what could happen in the near future, and react early to stay alive. This difference is also mirrored in the way our NLA principle looks for an error-minimizing trajectory of brain states. We postulate that the brain trades with near-future states and seeks for a path that minimizes errors of these future states any moment in time. Looking ahead towards what will likely happen allows the network for correcting the internal trajectory of deep neurons early enough so that the delayed output moves along the desired path. The notion of looking into the future to gate a dynamical system is also central in optimal control theory (as expressed by the Bellman equation, see e.g. Todorov, 2006). Yet, starting with a neuronal action is more principled as it includes the derivation of the dynamical system itself that will be optimally controlled.

The insight into the time structure of biological information processing allows us to express a simple form of a total ‘mismatch energy’ for our cortical neuronal networks, from which we derive the dynamic neuronal and synaptic laws. In short, the mismatch energy within a single pyramidal neuron is the squared prediction error between basal dendrites and the soma, together with the apical dendrites receiving a top-down feedback. The apical dendrites calculate a local prospective prediction error that looks ahead in time and overcomes neuronal integration delays (Fig. 1a). As a consequence, the output neurons are corrected on the fly by the prospective error processing, pushing them in real time closer to the desired path. In addition, the prospective errors are suited for gradient learning of the sensory synapses on the basal dendrites. This gradient learning is proven to reduce the error in the output neurons at any moment in time.

Somato-dendritic mismatch energies and the neuronal least-action (NLA) principle.

(a1) Sketch of a cross-cortical network of pyramidal neurons described by NLA. (a2) Correspondence between elements of NLA and biological observables such as membrane voltages and synaptic weights. (b1) The NLA principle postulates that small variations δũ (dashed) of the trajectories ũ (solid) leave the action invariant, δA = 0. It is formulated in the look-ahead coordinates ũ (symbolized by the spyglass) in which ‘hills’ of the Lagrangian (shaded grey zones) are foreseen by the prospective voltage so that the trajectory can turn by early enough to surround them. (b2) In the absence of output nudging (β = 0), the trajectory u(t) is solely driven by the sensory input, and prediction errors and energies vanish (L = 0, outer blue trajectory at bottom). When nudging the output neurons towards a target voltage (β > 0), somato-dendritic prediction errors appear, the energy increases (red dashed arrows symbolising the growing ‘volcano’) and the trajectory u(t) moves out of the L = 0 hyperplane, riding on top of the ‘volcano’ (red trajectory). Synaptic plasticity reduces the somato-dendritic mismatch along the trajectory by optimally ‘shoveling down the volcano’ (blue dashed arrows) while the trajectory settles in a new place on the L = 0 hyperplane (inner blue trajectory at bottom).

The NLA principle builds on and integrates various ingredients from existing work and theories. Output neurons, be they motor neurons or decision making neurons, are postulated to be ‘nudged’ towards the desired target time course by additional synaptic input to the soma or the proximal apical dendrite, as described by (Urbanczik & Senn, 2014). The cortical microcircuit with lateral ‘inhibition’ that seeks to cancel the top-down feedback in order to extract the apical error is inspired from Sacramento et al., 2018 and Haider et al., 2021. The energy-based approach for describing error-backpropagation for weak nudging is borrowed from the Equilibrium Propagation algorithm (Scellier & Bengio, 2017) that we generalize from a steady-state algorithm to real-time computation in cross-cortical microcircuits. Our theory covers both cases of weak and strong output nudging. For strong nudging, it likewise generalises the least-control principle (Meulemans, Zucchet, et al., 2022) and the prospective configuration algorithm (Song et al., 2024) from a steady-state to a dynamic real-time version, linking to optimal feedback control (Todorov & Jordan, 2002). Finally, the apical activity of our pyramidal neurons can be seen in the tradition of predictive coding (Rao & Ballard, 1999), where cortical feedback connections try to explain away lower-level activities. Yet, different from classical predictive coding, our prediction errors are integrated in the soma, and these errors are prospective in time. The errors extrapolate from current to future activities, so that their integration improves the network output in real time. The combination of an energy-based model with prospective coding in which neuronal integration delays are compensated on the fly enters also in Haider et al., 2021.

The paper is organized as follows: we first define the prospective somato-dendritic mismatch error, construct out of this the mismatch energy of a network, and ‘minimise’ this energy to obtain the error-corrected, prospective voltage dynamics of the network neurons. We then show that the prospective error coding leads to an instantaneous and joint processing of low-pass filtered input signals and backpropagated errors. Applied to motor control, the instantaneous processing is interpreted as a moving equilibrium hypothesis according to which sensory inputs, network state, motor commands and muscle feedback are in a self-consistent equilibrium at any point of the movement. We then derive a local learning rule that globally minimizes the somato-dendritic mismatch errors across the network, and show how this learning can be implemented through error-extracting cortical microcircuits and dendritic predictive plasticity.

Results

Somato-dendritic mismatch errors and the Lagrangian of cortical circuits

We consider a network of neurons – identified as pyramidal cells – with firing rates ri(t) in continuous time t. The somatic voltage ui of pyramidal neuron i is driven by the close-by basal input current, Σj Wijrj, with presynaptic rates rj and synaptic weights Wij, and an additional distal apical input ei that will be learned to represent a prospective prediction error at any moment in time (Fig. 1a). While in classical rate-based neuron models the firing rate ri of a neuron is a function of the somatic voltage, ρ(ui), the NLA principle implies that the effective firing rate of a cortical neuron is prospective. More concretely, the formalism derives a firing rate that linearly extrapolates from ρ(ui) into the future with the temporal derivative, , where represents the temporal derivative of ρ(ui(t)). There is experimental evidence for such prospective coding in cortical pyramidal neurons where the instantaneous rate ri is in fact not only a function of the underlying voltage, but also a function of how quickly that voltage increases (see Fig. 2a).

Prospective coding in cortical pyramidal neurons enables instantaneous voltage-to-voltage transfer.

(a1) The instantaneous spike rate of cortical pyramidal neurons (top) in response to sinusoidally modulated noisy input current (bottom) is phase-advanced with respect to the input (adapted from Köndgen et al., 2008). (a2) Similiarly, in NLA, the instantaneous firing rate of a model neuron , black) is phase-advanced with respect to the underlying voltage (u, red, postulating that the low-pass filtered rate is a function of the voltage, ). (b) Dendritic input in the apical tree (here called ē) is instantaneously causing a somatic voltage modulation (u, (modeling data from Ulrich, 2002)). The low-pass filtering with τ along the dendritic shaft is compensated by a lookahead mechanism in the dendrite . In Ulrich (2002) a phase advance is observed even with respect to the dendritic input current, not only the dendritic voltage, although only for slow modulations (as here). (c) While the voltage of the first neuron (u1) integrates the input rates rin from the past (bottom black upward arrows), the output rate r1 of that first neuron looks ahead in time, (red dashed arrows pointing into the future). The voltage of the second neuron (u2) integrates the prospective rates r1 (top black upwards arrows). By doing so, it inverts the lookahead operation, resulting in an instantaneous transfer from u1(t) to u2(t) (blue arrow and circles).

The second central notion of the theory is the prospective error ei, that we interpret as prospective somato-dendritic mismatch error in the individual network neurons, . It is defined as a mismatch between the prospective voltage, , and the weighted prospective input rates, Σj Wijrj. In the same way as the firing rates rj linearly extrapolate into the future given the current voltages uj of the presynaptic neurons j, the postsynaptic error is based on the linear extrapolation of its current voltage ui using its temporal derivative, . If the prospective error ei is low-pass filtered with time constant τ, it takes the form , where is the corresponding low-pass filtered firing rate of the presynaptic neuron j (that becomes a function of the presynaptic voltage, , see Methods, Sect. A). We refer to ēi as somato-dendritic mismatch error of neuron i that, as compared to ei, is non-prospective and instantaneous.

We next interpret the mismatch error ēi in terms of the morphology and biophysics of pyramidal neurons with basal and apical dendrites. While the error ei is formed in the apical dendrite, this error is low-pass filtered and added to the somatic voltage ui, that is also driven by the low-pass filtered basal input , so that . From the perspective of the basal dendrites, the low-pass filtered apical error ēi can be calculated as difference between the somatic voltage and the own local low-pass filtered input, . The somatic voltage ui is assumed to be sampled in the basal dendrite by the backpropagating acting potentials (Urbanczik & Senn, 2014; Spicher et al., 2017). The apical error now appears as ‘somato-basal’ mismatch error, that both are summarized as somato-dendritic mismatch error. It tells the difference between ‘what a neuron does’, which is based on the somatic voltage ui, and ‘what the basal inputs think it should do’, which is based on its own input (Fig. 1a2). The two quantities may deviate because neuron i gets additional ‘unpredicted’ apical inputs from higher-area neurons that integrates in the somatic voltage ui. What cannot be predicted in ui by the sensory-driven basal input remains as somato-basal (somato-dendritic) mismatch error ēi.

Associated with this mismatch error is the somato-dendritic mismatch energy defined for each network neuron i∈ 𝒩 as the squared mismatch error, On a subset of output neurons of the whole network, 𝒪 ⊆ 𝒩, a cost is defined as a function of the somatic voltage and some instructive reference signal such as targets or a reward. When a target trajectory is available, the cost is defined at each time point as a squared target error, Much more general mismatch energies and cost functions are conceivable, encompassing e.g. conductance-based neurons or including further dynamic variables (see SI, Sect. F). The cost represents a performance measure for the entire network that produces the output voltages uo(t) in response to some input rates rin(t). The cost directly relates to behavioral or cognitive measures such as the ability of an animal or human to perform a particular task in real time. The target could be provided by explicit external supervision, for example target movements in time encoded by , it could represent an expected reward signal, or it could arise via self-supervision from other internal prediction errors.

We define the Lagrangian (or ‘total energy’) of the network as a sum across all mismatch energies and costs, weighted by the nudging strength β of the output neurons, The low-pass filtered presynaptic rates, , also encompass the external input neurons. While in classical energy-based approaches L is called the total energy, we call it the ‘Lagrangian’ because it will be integrated along real and virtual voltage trajectories as done in variational calculus (leading to the Euler-Lagrange equations, see below and SI, Sect. F). We ‘prospectively’ minimize L locally across a voltage trajectory, so that, as a consequence, the local synaptic plasticity for Wij will globally reduce the cost along the trajectory (Theorem 1 below).

Due to the prospective coding, the Lagrangian can be minimal at any moment in time while the network dynamics evolves. This is different from the classical predictive coding (Rao & Ballard, 1999) and energy-based approaches (Scellier & Bengio, 2017; Song et al., 2024), where a stimulus needs to be fixed in time while the network relaxes to a steady state, and only there the prediction error is minimized (see SI, Sect. C, comparison with Equilibrium Propagation).

The least-action principle expressed for prospective firing rates

Motivated by the prospective firing in pyramidal neurons, we postulate that cortical networks strive to look into the future to prevent instantaneous errors. Each neuron tries to move along a trajectory that minimizes its own mismatch error ēi across time (Fig. 1b1). The ‘neuronal currency’ with which each neuron ‘trades’ with others to choose its own error-minimizing trajectory is the future discounted membrane potential, The prospective voltages ũ are the ‘canonical coordinates’ entering the NLA principle, and in these prospective coordinates the overall network searches for a ‘least-action trajectory’. Since from ũ we can recover the instantaneous voltage via (see SI, Sect. B), we can replace u in the Lagrangian and obtain L as a function of our new prospective coordinates ũ and the ‘velocities’ , i.e. , where bold fonts represent vectors. Inspired by the least-action principle from physics, we define the neuronal action A as a time-integral of the Lagrangian, The NLA principle postulates that the trajectory ũ(t) keeps the action A stationary with respect to small variations δũ (Fig. 1b1). In other words, nature chooses a trajectory such that, when deviating a little bit from it, say by δũ, the value of A will not change (or at most up to second order in the variation), formally δA = 0. The motivation to search for a trajectory that keeps the action stationary is borrowed from physics. The motivation to search for a stationary trajectory by varying the near-future voltages ũ, instead of u, is assigned to the evolutionary pressure in biology to ‘think ahead of time’. To not react too late, internal delays involved in the integration of external feedback need to be considered and eventually need to be overcome. In fact, only for the ‘prospective coordinates’ defined by looking ahead into the future, even when only virtually, will a real-time learning from feedback errors become possible (as expressed by our Theorems below).

The equations of motion that keep the action stationary with respect to these prospective coordinates are known to satisfy the Euler-Lagrange equations Applying these equations to our Lagrangian yields a prospective version of the classical leaky integrator voltage dynamics, with rates r and errors e that are looking into the future (Methods, Sects A, B), The ‘·’ denotes the component-wise product, and the weight matrix splits into weights from input neurons and weights from network neurons, W = (Win, Wnet). While for output neurons a target error can be defined, , for non-output neurons i no target exist and we hence set . In a control theoretic framework, the neuronal dynamics (Eq. 7a) represents the state trajectory, and the adjoint error dynamics (Eq. 7b) represents the integrated costate trajectory (Todorov, 2006).

From the point of view of theoretical physics, where the laws of motion derived from the least-action principle contain an acceleration term (as in Newton’s law of motion, like for a harmonic oscillator), one may wonder why no second-order time derivative appears in the NLA dynamics. As an intuitive example, consider driving into a bend. Looking ahead in time helps us to reduce the lateral acceleration by braking early enough, as opposed to braking only when the lateral acceleration is already present. This intuition is captured by minimizing the neuronal action A with respect to the discounted future voltages ũi instead of the instantaneous voltages ui. Keeping up an internal equilibrium in the presence of a changing environment requires to look ahead and compensate early for the predicted perturbations. Technically, the acceleration disappears because the Euler-Lagrange operator (Eq. 6) turns into a lookahead-gradient operator, , since the is absorbed via (see Methods, Sect. A, and SI, Sect. F for the link to the least-action principle in physics).

Mathematically, the voltage dynamics in Eq. 7a specifies an implicit differential equation since also appears on the right-hand side. This is because the prospective rates include through . Likewise, the prospective errors , with ē given in Eq. 7b and plugged into Eq. 7a, imply through . Nevertheless, the voltage dynamics can be stably run by replacing (t) on the right-hand side of Eq. 7a with the temporal derivative from the previous time step (technically, the Hessian is required to be strictly positive definite, see Methods Sect. F and SI, Sect. C). This ensures that the voltage dynamics of Eq. 7 can be implemented in cortical neurons with a prospective firing and a prospective dendritic error (see Fig. 2).

The error expression in Eq. 7b is reminiscent of error backpropagation (Rumelhart et al., 1986) and can in fact be related (Methods, Sect. C). Formally, the errors are backpropagated via transposed network matrix, , modulated by , the derivative of with respect to the underlying voltage. While the transpose can be constructed with various local methods (see Akrout et al., 2019; Max et al., 2022) in our simulations we mainly adhere to the phenomenon of feedback alignment (Lillicrap, Cownden, et al., 2016) and consider fixed and randomized feedback weights B (unless stated differently). Recent control theoretical work is exploiting the same prospective coding technique as expressed in Eq. 7 to tackle general time-varying optimization problems (see Simonetto et al., 2020 for a review and the SI, Sect. C for the detailed connection).

Prospective coding in neurons and instantaneous propagation

The prospective rates and errors entering via r and e in the NLA (Eq. 7) are consistent with the prospective coding observed in cortical pyramidal neurons in vitro (Köndgen et al., 2008). Upon sinusoidal current injection into the soma, the somatic firing rate is advanced with respect to its voltage (Fig. 2a), effectively compensating for the delay caused by the current integration. Likewise, sinusoidal current injection in the apical tree causes a lag-less voltage response in the soma (Fig. 2b, Ulrich, 2002). While the rates and errors in general can be reconstructed from their low-pass filterings via and , they become prospective in time because and ē are themselves instantaneous functions of the voltage u, and hence r and e depend on . The derivative of the membrane potential implicitly also appears in the firing mechanism of Hodgkin-Huxley-type conductances, with a quick depolarization leading to a stronger sodium influx due to the dynamics of the gating variables (Hodgkin & Huxley, 1952). This advances the action potential as compared to a firing that would only depend on u, not , giving an intuition of how such a prospective coding may arise. A similar prospective coding has been observed for retinal ganglion cells (Palmer et al., 2015) and cerebellar Purkinje cells (Ostojic et al., 2015), making a link from the visual input to the motor control.

To understand the instantaneous propagation through the network, we low-pass filter the dynamic equation (obtained by rearranging Eq. 7a), with ē given by Eq. 7b, to obtain the somatic voltage . At any point in time, the voltage is in a moving equilibrium between forward and backpropagating inputs. Independently of the network architecture, whether recurrent or not, the output is an instantaneous function of the low-pass filtered input and a putative correction towards the target, , see Fig. 2c and Methods, Sect. D. The mapping again expresses an instantaneous propagation of voltages throughout the network in response to both, the low-pass filtered input and feedback error . This instantaneity is independent of the network size, and in a feed-forward network is independent of its depths (see also Haider et al., 2021, where the instantaneity is on the rates, not the voltages). In the absence of the look-ahead activity, each additional layer would slow down the network relaxation time.

Notice that an algorithmic implementation of the time-continuous dynamics of a N -layer feedforward network would still need N calculation steps until information from layer 1 reaches layer N. However, this does not imply that an analog implementation of the prospective dynamics will encounter delays. To see why, consider a finite step-change Δu1 in the voltage of layer 1. In the absence of the look-ahead, Δu1 were mapped within the infinitesimal time interval dt to an infinitesimal change du2 in the voltages of layer 2. But with a prospective firing rate, , a step-change Δu1 translates to a delta-function in r1, this in turn to a step-change in the low-pass filtered rates , and therefore within dt to a step-change Δu2 in the voltages u2 of the postsynaptic neurons (Fig. 2c). Iterating this argument, a step-change Δu1 propagates ‘instantaneously’ through N layers within the ‘infinitesimal’ time interval N dt to a step-change ΔuN in the last layer. When run in a biophysical device in continuous time that exactly implements the dynamical equations (Eq. 7), the implementation becomes an instantaneous computation (since dt → 0). Yet, in a biophysical device, information has to be moved across space. This typically introduces further propagation delays that may not be captured in our formalism where low-pass filtering and prospective coding cancel each other exactly. Nevertheless, analog computation in continuous time, as formalized here, offers an idea to ‘instantaneously’ realize an otherwise time consuming numerical recipe run on time-discrete computing systems that operate with a finite clock cycle.

Prospective control and the moving equilibrium hypothesis

Crucially, at the level of the voltage dynamics (Eq. 7a) the correction is based on the prospective error e. This links our framework to optimal control theory and motor control where delays are also taken into account, so that a movement can be corrected early enough (Wolpert & Ghahramani, 2000; Todorov & Jordan, 2002; Todorov, 2004). The link between energy-based models and optimal control was recently drawn for strong nudging (β→ ∞) to learn individual equilibrium states (Meulemans, Zucchet, et al., 2022). Our prospective error e(t) appears as a ‘controller’ that, when looking at the output neurons, pushes the voltage trajectories towards the target trajectories. Depending on the nudging strength β, the control is tighter or weaker. For infinitely large β, the voltages of the output neurons are clamped to the time-dependent target voltages, (implying ), while their errors, , instantaneously correct all network neurons. For small β, the output voltages are only weakly controlled, and they are dominated by the forward input, .

To show how the NLA principle with the prospective coding globally maps to cortico-spinal circuits we consider the example of motor control. In the context of motor control, our network mapping can be seen as a forward internal model that quickly calculates an estimate of the future muscle length uo based on some motor plans, sensory inputs, and the current proprioceptive feedback (Fig. 3a). Forward models help to overcome delays in the execution of the motor plan by predicting the outcome, so that the intended motor plans and commands can be corrected on the fly (Kawato, 1999; Wolpert & Ghahramani, 2000).

Moving equilibrium hypothesis for motor control and real-time learning of cortical activity.

(a) A voluntary movement trajectory can be specified by the target length of the muscles in time, , encoded through the γ-innervation of muscle spindles, and the deviation of the effective muscle lengths from the target, . The Ia-afferents emerging from the spindles prospectively encode the error, so that their low-pass filtering is roughly proportional to the length deviation, truncated at zero (red). The moving equilibrium hypothesis states that the low-pass filtered input , composed of the movement plan and the sensory input (here encoding the state of the plant e.g. through visual and proprioceptive input, and ), together with the low-pass filtered error feedback from the spindles, , instantaneously generate the muscle lengths, , and are thus at any point in time in an instantaneous equilibrium (defined by Eq. 7). (b1) Intracortical iEEG activity recorded from 56 deep electrodes and projected to the brain surface. Red nodes symbolize the 56 iEEG recording sites modeled alternately as input or output neurons, and blue nodes symbolize the 40 ‘hidden’ neurons for which no data is available, but used to reproduce the iEEG activity. (b2) Corresponding NLA network. During training, the voltages of the output neurons were nudged by the iEEG targets (black input arrows, but for all red output neurons). During testing, nudging was removed for 14 out of these 56 neurons (here, represented by neurons 1, 2, 3). (c1) Voltage traces for the 3 example neurons in a2, before (blue) and after (red) training, overlaid with their iEEG target traces (grey). (c2) Total cost, integrated over a window of 8 s of the 56 output nodes during training with sequences of the same duration. The cost for the test sequences was evaluated on a 8 s window not used during training.

The observation that muscle spindles prospectively encode the muscle length and velocity (Dimitriou & Edin, 2010)) suggests that the prospective coding in the internal forward model mirrors the prospective coding in the effective forward pathway. This forward pathway leads from the motor plan to the spindle feedback, integrating also cerebellar and brainstem feedback (Kawato, 1999). Based on the motor plans, the intended spindle lengths and the effective muscle innervation are communicated via descending pathway to activate the γ- and α-motoneurons, respectively (Li et al., 2015). The mapping from the intended arm trajectory to the intended spindle lengths via γ-innervation is mainly determined by the joint geometry. The mapping from the intended arm trajectory to the force generating α-innervation, however, needs to also take account of the internal and external forces, and this is engaging our network W.

When we prepare an arm movement, spindles in antagonistic muscles pairs that measure the muscle length are tightened or relaxed before the movement starts (Papaioannou & Dimitriou, 2021). According to the classical equilibrium-point hypothesis (Feldman & Levin, 2009; Latash, 2010), top-down input adjusts the activation threshold of the spindles through (γ-)innervation from the spinal cord so that slight deviations from the equilibrium position can be signaled (Fig. 3a). We postulate that this γ-innervation acts also during the movement, setting an instantaneous target for the spindle lengths. The effective lengths of the muscle spindles is uo, and the spindles are prospectively signaling back the deviation from the target through the Ia-afferents (Dimitriou & Edin, 2010; Dimitriou, 2022). The low-pass filtered Ia-afferents may be approximated by a threshold-nonlinearity, , with β being interpreted as spindle gain (Latash, 2018). Combining the feedback from agonistic and antagonistic muscle pairs allows for extracting the scaled target error . Taking account of the prospective feedback, we postulate the moving equilibrium hypothesis according to which the instructional inputs, , the spindle feedback, , and the muscle lengths, uo, are at any point of the movement in a dynamic equilibrium. The moving equilibrium hypothesis extends the classical equilibrium-point hypothesis from the spatial to the temporal domain (for a formal definition of a moving equilibrium see Methods, Sect. F).

Prediction errors are also reduced when motor units within a muscle are recruited according to the size principle (Senn, Wyler, Clamann, Kleinle, Lüscher, et al., 1997), which itself was interpreted in terms of the physical least-action principle (Senn, Wyler, Clamann, Kleinle, Larkum, et al., 1995). With regard to the interpretation of the prospective feedback error as spindle activity, it is worth noticing that in humans the spindle activity is not only ahead of the muscle activation (Dimitriou & Edin, 2010), but also shares the property of a motor error (Dimitriou, 2016). The experiments show that during the learning of a gated hand movement, spindle activity is initially stronger when making movement errors, and it returns back to baseline with the success of learning. This observation is consistent with the NLA principle, telling that the proprioceptive prediction errors are minimized through the movement learning. We next address how the synaptic strengths W involved in producing the muscle length can be optimally adapted to capture this learning.

Local plasticity at basal synapses minimizes the global cost in real time

The general learning paradigm starts with input time series rin(t),i and target time series , while assuming that the target series are an instantaneous function of the low-pass filtered input series, . The low-pass filtering in the individual inputs could be with respect to any time constant (that may also be learned, see SI, Sect. B). Yet, for simplicity we assume the same time constant τ for low-pass filtering the rates of the network neurons and input neurons. The goal of learning is to adapt the synaptic strengths W in the student network so that this moves towards the target mapping, FW → F *. The local synaptic plasticity will also reduce the global cost C defined on the output neurons o in terms of the deviation of the voltage from the target, (Eq. 2).

The problem of changing synaptic weights to correct the behavior of downstream neurons, potentially multiple synapses away, is typically referred to as the credit assignment problem and is notoriously challenging in physical or biological substrates operating in continuous time. A core aspect of the NLA principle is how it relates the global cost C to the Lagrangian L and eventually to somato-dendritic prediction errors ē that can be reduced through local synaptic plasticity . We define this synaptic plasticity as partial derivative of the Lagrangian with respect to the weights, . Since the somato-dendritic mismatch error is , this leads to the local learning rule of the form ‘postsynaptic error times low-pass filtered presynaptic rate’, The plasticity rule runs simultaneously to the neuronal dynamics in the presence of a given nudging strength β that tells how strongly the voltage of an output neurons is pushed towards the target, . The learning rule is local in space since is represented as voltage of the basal dendrites, and the somatic voltage u may be read out at the synaptic site on the basal dendrite from the backpropagating action potentials that sample u at a given time (Urbanczik & Senn, 2014). The basal voltage becomes the dendritic prediction of the somatic activity u, interpreting Eq. 8 as ‘dendritic predictive plasticity’.

We have derived the neuronal dynamics as a path that keeps the action stationary. Without external teaching signal, the errors vanish and the voltage trajectory wriggles on the bottom of the energy landscape (L = 0, Fig. 1b2). If the external nudging is turned on, β > 0, errors emerge and hills grow out of the landscape. The trajectory still tries to locally minimize the action, but it is lifted upwards on the hills (L > 0, Fig. 1b2). Synaptic plasticity reshapes the landscape so that, while keeping β fixed, the errors are reduced and the landscape again flattens. The transformed trajectory settles anew in another place (inside the ‘volcano’ in Fig. 1b2). Formally, the local plasticity rule (Eq. 8) is shown to perform gradient descent on the Lagrangian and hence on the action. In the energy landscape picture, plasticity ‘shovels off’ energy along the voltage path so that this is lowered most efficiently. The error that is back-propagated through the network tells at any point on the voltage trajectory how much to ‘dig’ in each direction, i.e. how to adapt the basal input in each neuron in order to optimally lower the local error.

The following theorem tells that synaptic plasticity pushes the network mapping towards the target mapping at any moment in time. The convergence of the mapping is a consequence of the fact the plasticity reduces the Lagrangian L = EM + βC along its gradient.

(real-time Dendritic Error Propagation, rt-DeEP)

Consider an arbitrary network W with voltage and error dynamics following Eq. 7. Then the local plasticity rule (Eq. 8), acting at each moment along the voltage trajectories, is gradient descen

on the Lagrangian L for any nudging strength β ≥ 0, i.e. , with .

on the cost C for small nudging, β → 0, while up-scaling the error to , i.e. .

The gradient statements hold at any point in time (long enough after initialization), even if the input trajectories rin(t) contain delta-functions and the target trajectories contain step-functions.

Loosely speaking, the NLA enables the network to localize in space and time an otherwise global problem: what is good for a single neuron (the local plasticity) becomes good for the entire network (the gradient on the global cost). Learning is possible any point in time along the trajectory because the NLA inferred a prospective voltage dynamics expressed in prospective firing rates ri and prospective errors ei of the network neurons. In the limit of strong nudging (β → ∞), the learning rule performs gradient descent on the mismatch energies in the individual neurons. If the network architecture is powerful enough so that after learning all the mismatch energies vanish, , then the cost will also vanish,. This is because for the output neurons the mismatch error includes the target error (Eq. 7b). In the limit of weak nudging (β 0), the learning rule performs gradient descent on C, and with this also finds a local minimum of the mismatch energies.

In the case of weak nudging and a single steady-state equilibrium, the NLA algorithm reduces to the Equilibrium Propagation algorithm (Scellier & Bengio, 2017) that minimizes the cost C for a constant input and a constant target. In the case of strong nudging and a single steady-state equilibrium, the NLA principle reduces to the Least-Control Principle (Meulemans, Zucchet, et al., 2022) that minimizes the mismatch energy EM for a constant input and a constant target, with the apical prediction error becoming the prediction error from standard predictive coding (Rao & Ballard, 1999). While in the Least-Control Principle the inputs and outputs are clamped to fixed values, the output errors are backpropagated and the network equilibrates in a steady state where the corrected network activities reproduce the clamped output activities. This state is called the ‘prospective configuration’ in Song et al., 2024 because neurons deep in the network are informed about the distal target already during the inference, and are correspondingly adapted to be consistent with this distal target. In the NLA principle, after an initial transient, the network always remains in the moving equilibrium due to the prospective coding. While inputs and targets dynamically change, the network moves along a continuous sequence of prospective configurations.

In the motor control example, the theorem tells that a given target motor trajectory is learned to be reproduced with the forward model , by applying the dendritic predictive plasticity for the network neurons (Eq. 8). We next exemplify the theory by looking into the brain, reproducing cortical activity, and showing how a multi-layer cortical network can learn a sensory-motor mapping while staying in a moving equilibrium throughout the training.

Reproducing intracortical EEG recordings and recognizing handwritten digits

As illustration we consider a recurrently connected network that learns to represent intracortical electroencephalogram (iEEG) data from epileptic patients (Fig. 3b). For each electrode, we assign a neuron within this network to represent the activity of the cell cluster recorded in the corresponding iEEG signal via its membrane potential. During learning, a randomly selected subset of electrode neurons are nudged towards the target activity from recorded data while learned to be reproduced by the other neurons. After learning, we can present only a subset of electrode neurons with previously unseen recordings and observe how the activity of the other neurons closely matches the recordings of their respective electrodes (Fig. 3c). The network derived from NLA is thus able to learn complex correlations between signals evolving in real time by embedding them in a recurrent connectivity structure.

As an example of sensory-motor processing in the NLA framework, we next consider a well-studied image recognition task, here reformulated in a challenging time-continuous setting, and interpreted as motor task where 1 out of 10 fingers has to be bent upon seeing a corresponding visual stimulus (see Fig. 3a). In the context of our moving equilibrium hypothesis, we postulate that during the learning phase, but not the testing phase, an auditory signal identifies the correct finger and sets the target spindle lengths of 10 finger flexors, . The target spindle length encodes the desired contraction of a flexor muscle in the correct finger upon the visual input rin(t), and a corresponding relaxation for the 9 incorrect fingers.

We train a hierarchical three-layer network on images of handwritten digits (MNIST, LeCun, 1998), with image presentation times between 0.5τ (= 5 ms) and 20τ (= 200 ms, with τ = 10 the membrane time constant). Fig. 4a-c1 depict the most challenging scenario with the shortest presentation time. Synaptic plasticity is continuously active, despite the network never reaching a temporal steady state (Fig. 4b1). Due to the lookahead firing rates in the NLA, the mismatch errors ēi(t) represent the correct gradient and propagate without lag throughout the network. As a consequence, our mismatch errors are almost equal to the errors obtained from classical error backpropagation applied at each time step to the purely forward network (i.e. the network that suppresses the error-correction ē of the voltage and instead considers the ‘classical’ voltage ul = Wl ρ(ul−1) only, see blue dots in Fig. 4b2). The network eventually learned to implement the mapping with a performance comparable to error-backpropagation at each dt, despite the short presentation time of only 5 ms (Fig. 4c1). The approximation is due to the fact that the NLA learns an instantaneous mapping from the low-pass filtered input rates to the output voltage uo, while the mapping from the original input rates rin to the voltages u1 of the first-layer neurons (and hence also to the output voltages uo) is delayed by τin. Since in the simulations the target voltages were switched instantaneously with rin (and not with ), however, a mismatch error between uo and remains for stimulus presentation times shorter than τin (Fig. 4c2). The Latent Equilibrium (Haider et al., 2021) avoids these temporal limitations by implementing an instantaneous mapping on the rates instead on the voltages (Methods, Sect. F).

On-the-fly learning of finger responses to visual input with real-time Dendritic Error Propagation (rt-DeEP).

(a) Functionally feedforward network with handwritten digits as visual input in Fig. 3a, here from the MNIST data set, 5 ms presentation time per image), backprojections enabling credit assignment, and activity of the 10 output neurons interpreted as commands for the 10 fingers (forward architecture: 784 × 500 ×10 neurons). (b) Example voltage trace (b1) and local error (b2) of a hidden neuron in NLA (red) compared to an equivalent network without lookahead rates (orange). Note that neither network achieves a steady state due to the extremely short input presentation times. Errors calculated via exact backpropagation, i.e. by using the error backpropagation algorithm on a purely feedforward NLA network at every simulation time step (with output errors scaled by β), shown for comparison (blue dots). (c) Comparison of network models during and after learning. Color scheme as in (b). (c1) The test error under NLA evolves during learning on par with classical error backpropagation performed each Euler dt based on the feedforward activities. In contrast, networks without lookahead rates are incapable of learning such rapidly changing stimuli. (c2) With increasing presentation time, the performance under NLA further improves, while networks without lookahead rates stagnate at high error rates. This is caused by transient, but long-lasting misrepresentation of errors following stimulus switches: when plasticity is turned off during transients and is only active in the steady state, comparably good performance can be achieved (dashed orange). (d) Receptive fields of 6 hidden-layer neurons after training, demonstrating that even for very brief image presentation times (5ms), the combined neuronal and synaptic dynamics are capable of learning useful feature extractors such as edge filters.

The instantaneous voltage propagation reduces an essential constraint of previous models of bio-plausible error backpropagation (e.g., Scellier & Bengio (2017), Whittington & Bogacz (2017), and Sacramento et al. (2018), with reviews Richards et al. (2019), Whittington & Bogacz (2019), and Lillicrap, Santoro, et al. (2020)): without lookahead firing rates, networks need much longer to correctly propagate errors across layers, with each layer roughly adding another membrane time constant of 10 ms, and thus cannot cope with realistic input presentation times. In fact, in networks without lookahead output, learning is only successful if plasticity is switched off while the network dynamics did not reach a stationary state during a stimulus presentation interval (Fig. 4c2). As a comparison, neuronal response latency to flashing stimuli in the primary visual cortex (V1) of rats in a synchronized state are in the order of 50 ms (Wang et al., 2014). The latency shortens to roughly 40 ms if in a desynchronized state (Wang et al., 2014).

In humans, cortical responses to expected visual stimuli (rectangulars) are observed 70-90 ms ahead of the responses observed in the absence of expectations (which have a latency of roughly 150 ms, see Blom et al., 2020). A remaining latency in the presence of prospective coding is consistent with a remaining response delay of τin in the presented NLA principle. The minimal cortical response latency of roughly 50 ms to an expected rectangular (Blom et al., 2020) compares with the roughly 80 ms cortical response time when humans have to decide whether a flashed image contains an animal or not (Thorpe et al., 1996). The ongoing higher cortical processing during animal detection does not seem to substantially slow down cortical responses to sensory stimuli. This supports a ‘sensory equilibrium hypothesis’, the sensory correlate of our ‘moving equilibrium hypothesis’ for motor control, that enables us to interact with a dynamic visual environment in almost real time (as also suggested in Blom et al., 2020). Notice that, from a technical perspective, making the time constants of individual cortical neurons arbitrarily short leads to network instabilities and is unlikely the option chosen by the brain (see SI Sect. C, comparison with Equilibrium Propagation).

Implementation in cortical microcircuits

So far, we did not specify how errors e appearing in the differential equation for the voltage (Eq. 7a) are transmitted across the network in a biologically plausible manner. Building on Sacramento et al., 2018, we propose a cortical microcircuit to enable this error transport, with all neuron dynamics evolving according to the NLA principle. Although the idea applies for arbitrarily connected networks, we use the simpler case of functionally feedforward networks to illustrate the flow of information in these microcircuits (Fig. 5a).

Hierarchical plastic microcircuits implement real-time Dendritic Error Learning (rt-DeEL).

(a) Microcircuit with ‘top-down’ input (originating from peripheral motor activity, blue line) that is explained away by the lateral input via interneurons (dark red), with the remaining activity representing the error ēl. Plastic connections are denoted with a small red arrow and nudging with a dashed line. (b1) Simulated network with 784-300-10 pyramidal-neurons and a population of 40 interneurons in the hidden layer used for the MNIST learning task where the handwritten digits have to be associated to the 10 fingers. (b2) Test errors for rt-DeEL with joint tabula rasa learning of the forward and lateral weights of the microcircuit. A similar performance is reached as with classical error backpropagation. For comparability, we also show the performance of a shallow network (dashed line). (b3) Angle derived from the Frobenius norm between the lateral pathway and the feedback pathway BlWl+1. During training, both pathways align to allow correct credit assignment throughout the network. Indices are dropped in the axis label for readability.

For such an architecture, pyramidal neurons in area l (that is a ‘layer’ of the feedforward network) are accompanied by a pool of interneurons in the same layer (area). The dendrites of the interneurons integrate in time (with time constant τ) lateral input from pyramidal neurons of the same layer (rl) through plastic weights . Additionally, interneurons receive ‘top-down nudging’ from pyramidal neurons in the next layer through randomly initialized and fixed backprojecting synapses targeting the somatic region, and interneuron nudging strength βI. The notion of ‘top-down’ originates from the functionally feed-forward architecture leading from sensory to ‘higher cortical areas’. In the context of motor control, the highest ‘area’ is last stage controlling the muscle lengths, being at the same time the first stage for the proprioceptive input (Fig. 3a).

According to the biophysics of the interneuron, the somatic membrane potential becomes a convex combination of the two types of afferent input (Urbanczik & Senn, 2014), In the biological implementation, the feedback input is mediated by the low-pass filtered firing rates , not by ul+1 as expressed in the above equation. Yet, we argue that for a threshold-linear ρ the ‘top-down nudging’ by the rate is effectively reduced to a nudging by the voltage ul+1. This is because errors are only backpropagated when the slope of the transfer function is positive, , and hence when the upper-layer voltage is in the linear regime. For more general transfer functions, we argue that short-term synaptic depression may invert the low-pass filtered presynaptic rate back to the presynaptic membrane potential, , provided that the recovery time constant τ matches the membrane time constant (see end of Results and SI, Sect. A).

Apical dendrites of pyramidal neurons in each layer receive top-down input from the pyramidal population in the upper layer through synaptic weights Bl. These top-down weights could be learned to predict the lower-layer activity (Rao & Ballard, 1999) or to become the transposed of the forward weight matrix (, Max et al., 2022), but for simplicity we randomly initialized them and keep them fixed (Lillicrap, Santoro, et al., 2020). Beside the top-down projections the apical dendrites also receive lateral input via an interneuron population in the same layer through synaptic weights – that are plastic and will be learned to obtain suitable dendritic errors. The ‘-’ sign is suggestive for these interneurons to subtract away the top-down input entering through Bl (while the weights can still be positive or negative). Assuming again a conversion of rates to voltages, also for the inhibitory neurons that may operate in a linear regime, the overall apical voltage becomes What cannot be explained away from the top-down input Blul+1 by the lateral feedback, , remains as dendritic prediction error in the apical tree (Fig. 5a). If the top-down and lateral feedback weights are learned as outlined next, these apical prediction errors take the role of the backpropagated errors in the classical backprop algorithm.

To adjust the interneuron circuit in each layer (‘area’), the synaptic strengths from pyramidal-to-interneurons, , are learned to minimize the interneuron mismatch energy, . The interneurons, while being driven by the lateral inputs , learn to reproduce the upper-layer activity that also nudges the interneuron voltage. Learning is accomplished if the upper-layer activity, in the absence of an additional error on the upper layer, is fully reproduced in the interneurons by the lateral input.

Once the interneurons learned to represent the ‘error-free’ upper-layer activity, they can be used to explain away the top-down activities that also project to the apical trees. The synaptic strengths from the inter-to-pyramidal neurons, , are learned to minimize the apical mismatch energy, . While in the absence of an upper-layer error, the top-down activity Blul+1 can be fully cancelled by the interneuron activity , a neuron-specific error will remain in the apical dendrites of the lower-level pyramidal neurons if there was an error endowed in the upper-layer neurons. Gradient descent learning on these two energies results in the learning rules for the P-to-I and I-to-P synapses, The following theorem on dendritic error learning tells that the plasticity in the lateral feedback loop leads to an appropriate error representation in the apical dendrites of pyramidal neurons.

(real-time Dendritic Error Learning, rt-DeEL)

Consider a cortical microcircuit composed of pyramidal and interneurons, as illustrated in Fig. 5a, with more interneurons in layer (‘cortical area’) l than pyramidal neurons in layer l+1, and with adaptable pyramidal-to-inhibitory weights within the same layer (that are nudged through top-down weights , see Methods, Sect. E). Then, for suitable top-down nudging, learning rates, and initial conditions, the inhibitory-to-pyramidal synapses within each layer l (Eq. 11) evolve such that the lateral feedback circuit aligns with the bottom-up-top-down feedback circuit, After this horizontal-to-vertical circuit alignment, the apical voltages of the layer-l pyramidal neurons (Eq. 10) represent the ‘B-backpropagated’ errors. When modulated by the postsynaptic rate derivatives, , the apical voltages yield the appropriate error signals for learning the forward weights Wl according to (Eq. 8).

The backprojecting weights can also be learned by a local real-time learning rule to become transpose of the forward weights, (Max et al., 2022). In this case, the error signals ēl learned in the apical dendrites according to the above Theorem (Eq. 13) represent the gradient errors ē appearing in the real-time dendritic error propagation (rt-DeEP, Theorem 1). There, the errors ē drive the gradient plasticity of the general weight matrix W, split up here into the forward weights Wl to a layer l (for l = 1, .., N).

Simultaneously learning apical errors and basal signals

Microcircuits following these neuronal and synaptic dynamics are able to learn the classification of hand-written digits from the MNIST dataset while learning the apical signal representation (Fig. 5b1-2). In this case, feedforward weights Wl and lateral weights and are all adapted simultaneously. Including the -plasticity (by turning on the interneuron nudging from the upper layer, βI > 0 in Eq. 9), greatly speeds up the learning.

With and without -plasticity, the lateral feedback via interneurons (with effective weight ) learns to align with the forward-backward feedback via upper layer pyramidal neurons (with effective weight BlWl+1, Fig. 5b3). The microcircuit extracts the gradient-based errors (Eq. 13), while the forward weights use these errors to reduce these errors to first minimize the neuron-specific mismatch errors, and eventually the output cost.

Since the apical voltage appears as a postsynaptic factor in the plasticity rule for the interneurons , this I-to-P plasticity can be interpreted as Hebbian plasticity of inhbitory neurons, consistent with earlier suggestions (Vogels et al., 2012; Bannon et al., 2020). The plasticity of the P-to-I synapses, in the same way as the plasticity for the forward synapses , can be interpreted as learning from the dendritic prediction of somatic activity (Urbanczik & Senn, 2014).

Crucially, choosing a large enough interneuron population, the simultaneous learning of the lateral microcircuit and the forward network can be accomplished without fine-tuning of parameters. As an instance in case, all weights shared the same learning rate. Such stability bolsters the biophysical plausibility of our NLA framework and improves over the previous, more heuristic approach (Sacramento et al., 2018; Mesnard et al., 2019). The stability may be related to the nested gradient descent learning according to which somatic and apical mismatch errors in pyramidal neurons, and somatic mismatch errors in inhibitory neurons are minimized.

Finally, since errors are defined at the level of membrane voltages (Eq. 11), synapses need a mechanism by which they can recover the presynaptic voltage errors from their afferent firing rates. While for threshold-linear transfer functions the backpropagated voltage errors translate into rate errors (SI, Sect. A), more general neuronal nonlinearities must be matched by corresponding synaptic nonlinearities. Pfister et al. (2010) have illustrated how spiking neurons can leverage short-term synaptic depression to estimate the membrane potential of their presynaptic partners. Here, we assume a similar mechanism in the context of our rate-based neurons. The monotonically increasing neuronal activation function, , can be approximately compensated by a vesicle release probability that monotonically decreases with the low-pass filtered presynaptic rate (see SI, Sect. A and Fig. 6 therein). If properly matched, this leads to a linear relationship between the presynaptic membrane potential ul+1 and the postsynaptic voltage contribution.

Recovering presynaptic potentials through short term depression.

(a1) Relative voltage response of a depressing cortical synapse (recreated from Abbott et al., 1997), identified as synaptic release probability p. (a2) The product of the low-pass filtered presynaptic firing rate times the synaptic release probability is proportional to the presynaptic membrane potential, . (a3) Average in vivo firing rate of a neuron in the visual cortex as a function of the somatic membrane potential (recreated from Anderson et al., 2000), which can be qualitatively identified as the stationary rate derived in Eq. 43.

Discussion

We introduced a least-action principle to neuroscience for deriving the basic laws of the voltage and synaptic dynamics in networks of cortical neurons. The approach is inspired by the corresponding principle in physics where basic laws of motions are derived across the various scales. While in physics the action is defined as time-integral of the kinetic minus potential energy, we define the action as time-integral of instantaneous somato-dendritic mismatch errors across network neurons plus a behavioural error. The ‘kinetics’ of a voltage trajectory only arises because we postulate that the action along a trajectory is minimized with respect to future voltages, not the instantaneous voltage, as would be done in physics. The postulate implies a prospective voltage dynamics that looks ahead in time, together with prospective local errors, in order to minimize the action and hence the somato-dendritic mismatch errors. The prospective errors nudge the firing of pyramidal neurons deep in the brain, so that motor neurons improve the output of the network right in time. A putative behavioural error, encoded in the motor feedback, propagates back through the network and produces prospective corrections of the pyramidal neuron activities that effectively manifest in instantaneous corrections of the motor trajectory. Through this prospective coding, the sensory stream, the deep network activity, and the motor feedback are in sync at any moment in time. We formulated the dynamic synchronization as ‘moving equilibrium hypothesis’, referring to the classical equilibrium point hypothesis for motor control (Feldman & Levin, 2009; Latash, 2010). More generally, the brain activity formed by the prospective firing of cortical pyramidal neurons is in a moving equilibrium while converting sensory input streams into motor outputs, consistent with prospective sensory processing in the human cortex (Blom et al., 2020).

Because the neuronal dynamics derived from the global NLA principle is in a moving equilibrium, the prospective dendritic errors that globally correct the output trajectory are also suited to instruct local synapatic plasticity in the dendrites. In fact, working down the apical errors by adapting the sensory driven synapses on the basal dendrites reduces the global output errors in real-time. The apical errors are extracted from the top-down feedback via lateral ‘inhibition’ that tries to cancel the top-down signal. This top-down feedback includes activity from a putative erroneous motor output that was not foreseen by the local inhibition and thus survives as a local apical error. Given the prospective coding of the pyramidal neurons, the dendritic errors are also prospective and thus able to induce the correct error-minimizing plasticity online, while stimuli and targets continuously change.

The NLA principle as a bottom-up theory from neurons to behaviour

To show that the NLA principle offers a viable program for a formalization of neuroscience following the example of physics, we exemplified its ramifications into dendritic computation, cortical microcircuits, synaptic plasticity, motor control, and sensory-based decision making. The crucial point of our axiomatization is that it connects the local neuronal errors to the global behavioural errors right in the formulation of the principle, eventually leading to local gradient-based plasticity rules. Because the formulation builds upon computations that can be realized in single neurons and dendrites to produce a behavioural output, the NLA principle can be seen as a bottom-up theory of behaviour. It is articulated in terms of apical and basal dendrites, somatic firing, network connectivity and behavioural outputs that jointly minimize their errors. This contrasts the related free energy principle, for instance, that leads to a top-down theory of behaviour by starting with the statistical, but more universal, notion of a free energy. It postulates that any self-organizing system, that is at a statistical equilibrium with its environment, must minimize its free energy (Friston, 2010; Friston et al., 2022), and from there works down its way to neurons and dendrites (Bastos et al., 2012; Kiebel & Friston, 2011).

Starting with a single Lagrangian function that specifies the form of the somato-dendritic prediction errors leaves some freedom for the interpretation and the implementation of the emerging dynamical equations for the voltages. We interpret errors to be represented in the apical dendrites of pyramidal neurons while sensory input targets the basal dendrites, but other dendritic configurations are conceivable (Mikulasch, Rudelt, Wibral, et al., 2023) that apply also to non-pyramidal neurons. We have chosen a specific interneuron circuitry to extract our apical errors, but other microcircuits or error representations might also be considered (Keller & Mrsic-Flogel, 2018). On the other hand, the derived gradient-based synaptic plasticity is tightly linked to the specific form of the somato-dendritic prediction errors expressed in the Lagrangian and its interpretation, making specific predictions for synaptic plasticity (as outlined below). The ‘external’ feedback entering through the cost function offers additional freedom to model behavioural interactions. We considered an explicit time course of a target voltage in motor neurons, for instance imposed by the feedback from muscle spindles that are themselves innervated by a prospective top-down signal to control muscle lengths (Papaioannou & Dimitriou, 2021; Dimitriou, 2022). But the cost may also link to reinforcement learning and express a delayed reward feedback delivered upon a behavioural decision of an agent acting in a changing environment (Friedrich, Urbanczik, et al., 2011; Friedrich & Senn, 2012).

A fundamental difficulty arises when the neuronal implementation of the Euler-Lagrange equations requires an additional microcircuit with its own dynamics. This is the case for the suggested microcircuit extracting the local errors. Formally, the representation of the apical feedback errors first needs to be learned before the errors can teach the feedforward synapses on the basal dendrites. We showed that this error learning can itself be formulated as minimizing an apical mismatch energy. What the lateral feedback through interneurons cannot explain away from the top-down feedback remains as apical prediction error. Ideally, while the network synapses targetting the basal tree are performing gradient descent on the global cost, the microcircuit synapses involved in the lateral feedback are performing gradient descent on local error functions, both at any moment in time. The simulations show that this intertwined system can in fact learn simultaneously with a common learning rate that is properly tuned. The cortical model network of inter- and pyramidal neurons learned to classify handwritten digits on the fly, with 10 digit samples presented per second. Yet, the overall learning is more robust if the error learning in the apical dendrites operates in phases without output teaching but with corresponding sensory activity, as may arise during sleep (see e.g. Deperrois et al., 2022; Deperrois et al., 2023).

The NLA principle integrates classical theories for cortical processing and learning

The prospective variational principle introduced with the NLA allows for integrating previous ideas on formalizing the processing and learning in cortical networks. Four such classical lines of theories come together. (i) The first line refers to the use of an energy function to jointly infer the neuronal dynamics and synaptic plasticity, originally formulated for discrete-time networks (Hopfield, 1982; Ackley et al., 1985), and recently extended to continuous-time networks (Scellier & Bengio, 2017). (ii) The second line refers to understanding error-backpropagation in the brain (Rumelhart et al., 1986; Xie & Seung, 2003; Whittington & Bogacz, 2017; Whittington & Bogacz, 2019; Lillicrap, Santoro, et al., 2020). (iii) The third line refers to dendritic computation and the use of dendritic compartmentalization for various functions such as nonlinear processing (Schiess et al., 2016; Poirazi & Papoutsi, 2020) and deep learning (Guerguiev et al., 2017; Sacramento et al., 2018; Haider et al., 2021). (iv) The fourth line refers to predictive coding (Rao & Ballard, 1999) and active inference (Pezzulo et al., 2022) to improve the sensory representation and motor output, respectively.

With regard to energy functions, the NLA principle adds a variational approach to characterize continuous-time neuronal trajectories and plasticity. Variational approaches are studied in the context of optimal control theory where a cost integral is minimized across time, constrained to some network dynamics (Todorov & Jordan, 2002; Meulemans, Farinha, et al., 2021). The NLA represents a unifying notion that allows to infer both, the network dynamics and its optimal control from a single Lagrangian. The error we derive represents prospective control variables that are applied to the voltages of each network neurons so that they push the output neurons towards their target trajectory. The full expression power of this control theoretic framework has yet to be proven when it is extended to genuine temporal processing that includes longer time constants, for instance inherent in a slow threshold adaptation (Bellec et al., 2020). The NLA principle can also treat the case of strong feedback studied so far in relaxation networks only (Meulemans, Zucchet, et al., 2022; Song et al., 2024). Our rt-DeEP Theorem makes a statement for real-time gradient descent learning while the network is in a moving equilibrium, linking to motor learning in the presence of perturbing force fields (Herzfeld et al., 2014) or perturbing visual inputs (Dimitriou, 2016).

With regard to error-backpropagation, the NLA principle infers a local error that originates from the error in the output neurons. In the current version, the NLA relies on feedback alignment (Lillicrap, Cownden, et al., 2016) to obtain a local apical errors without postulating plasticity in the top-down synapses. Other works have explored learning the feedback weights (Akrout et al., 2019; Kunin et al., 2020), notably within a phase-less real-time learning framework as considered here (Max et al., 2022). It would be promising to combine these ideas to obtain a fully plastic microcircuit that adjusts from scratch to various real-time learning tasks.

With regard to dendritic computation, the NLA principle extends the idea of dendritic error representation (Sacramento et al., 2018; Mesnard et al., 2019) by the prospective coding of both errors and firing rates (Fig. 2). As a consequence, the various dendritic delays are compensated and synaptic plasticity can operate at any moment, without need to wait for network relaxations (Scellier & Bengio, 2017; Song et al., 2024). In the present framework, the input rates are low-pass filtered by variable input time constants (τin) before the induced voltages are instantaneously processed in the network. While this offers the possibilities for a temporal processing of the inputs, step-changes in the input rates cannot instantaneously propagate (as made possible in Haider et al., 2021). The dendritic error representation has also been applied to spike-based learning in recurrent networks (Mikulasch, Rudelt & Priesemann, 2021).

With regard to the principle of predictive coding (Rao & Ballard, 1999; Pezzulo et al., 2022), the NLA offers a form of predictive error-coding in the apical tree of sensory pyramidal neurons that shapes the sensory representation. Both the predictive coding and the NLA principle imply an error-based adaptation of the lower and higher cortical representations. However, while predictive coding is based on propagating errors, the NLA principle directly propagates the error-corrected representations. Based on the backpropagated activities, it extracts in each area the prospective errors to reshape the local representations and induce synaptic plasticity. Otherwise, the freedom in defining the cost function in the NLA, and the inclusion of active inference in predictive coding (Pezzulo et al., 2022), makes the two frameworks equally powerful. Both frameworks can explain the learning from motor and sensory prediction errors. In fact, our example of motor control minimizes the proprioceptive error formed by the target muscle lengths minus the effective muscle lengths. Active inference likewise minimizes the mismatch between an internal motor target and the proprioceptive inputs caused by the own actions. In turn, the cost function in the NLA principle may also capture the sensory prediction errors, where the ground truth lies in the external inputs that are learned to be matched by the sensory expectation. Hence, while being functionally equivalent, the neuronal interpretations are different: predictive coding focuses on the propagation of errors using the sensory- and motor-representations as local auxiliary quantities, and the NLA principle directly integrates errors in ‘auxiliary dendrites’ to propagate sensory or motor representations.

The NLA integrates and predicts features of synapses, dendrites and circuits

Motivated by the predictive power of the least-action principle in physics, we ask about experimental confirmation and predictions of the NLA principle. Given its axiomatic approach, it appears astonishing to find various preliminary matches at the dendritic, somatic, interneuron, synaptic and even behavioural level. Some of these are:

the prospective coding of pyramidal neuron firing (Köndgen et al., 2008);

the prospective processing of apical signals while propagating to the soma (Ulrich, 2002);

the basal synaptic plasticity on pyramidal neurons and synaptic plasticity on interneurons, driven by the postsynaptic activity that is ‘unexplained’ by the distal dendritic voltage (Urbanczik & Senn, 2014) – while partly consistent with spike-timing dependent plasticity (Sjöström et al., 2001; Spicher et al., 2017), the postulated dendritic voltage-dependence of the plasticity rules still awaits an experimental inspection;

the Hebbian homeostatic plasticity of interneurons targeting the apical dendritic tree of pyramidal neurons (Bannon et al., 2020);

the short-term synaptic depression at top-down synapses targetting inhibitory neurons and apical dendrites (akin to Abbott et al., 1997, but with a faster recovery time constant) that invert the presynaptic activation function (see also Pfister et al., 2010);

the modulation of the apical contribution to the somatic voltage by the slope of the somatic activation function (for instance by downregulating apical NMDA receptors with increasing rate of backpropagating action potentials, Theis et al., 2018); and

the involvement of muscle spindles in the prospective encoding of motor errors during motor learning, with the γ-innervation setting the target length of the spindles (Dimitriou, 2016; Papaioannou & Dimitriou, 2021; Vargas et al., 2023).

More experimental and theoretical work is required to substantiate these links and test specific predictions, such as the apical error representation in cortical pyramidal neurons.

Overall, our approach adapts the least-action principle from physics to be applied to neuroscience, and couples it with a normative perspective on the prospective processing of neurons and synapses in global cortical networks and local microcircuits. Given its physical underpinnings, the approach may inspire the rebuilding of computational principles of cortical neurons and circuits in neuromorphic hardware (Bartolozzi et al., 2022). A step in this direction, building on the instantaneous computational capabilities with slowly integrating neurons, has been done in Haider et al., 2021. Given its aspiration for a theoretical framework in neurobiology, a next challenge would be to generalize the NLA principle to spiking neurons (Gerstner & Kistler, 2002; Brendel et al., 2020) with their potential for hardware implementation (Zenke & Ganguli, 2018; Göltz et al., 2021; Cramer et al., 2022), to include attentional mechanisms in terms of dendritic gain modulation (Larkum et al., 2004) with a putative link to self-attention in artificial intelligence (Vaswani et al., 2017), to add second-order errors to cope with certainties (Granier et al., 2023), and to incorporate longer temporal processing as, for instance, offered by neuronal adaptation processes (La Camera et al., 2006) or realistically modelled dendrites (Chavlis & Poirazi, 2021).

Methods

Euler-Lagrange equations as inverse low-pass filters

The theory is based on the lookahead of neuronal quantities. In general, the lookahead of a trajectory x(t) is defined via lookahead operator applied to x, The lookahead operator is the inverse of the low-pass filter operator denoted by a bar, This low-pass filtering can also be characterized by the differential equation , see SI, Sect.B. Hence, applying the low-pass filtering to x and then the lookahead operator to , and using the Leibnitz rule for differentiating an integral, we calculate . In turn, applying first the lookahead, and then the low-pass filtering, also yields the original trace back, .

We consider an arbitrary network architecture with network neurons that are recurrently connected and that receive external input through an overall weight matrix W = (Win, Wnet), aggregated column-wise. The instantaneous presnyaptic firing rates are r = (rin, rnet)T, interpreted as a single column vector. A subset of network neurons are output neurons, 𝒪 ⊆ 𝒩, for which target voltages u* may be imposed. Rates and voltages may change in time t. Network neurons are assigned a voltage u, generating the low-pass filtered rate , and a low-pass filtered error . We further define output errors for o ∈ O, and for non-output neurons i ∈ 𝒩 \ 𝒪. With this, the Lagrangian from Eq. 3 takes the form We next use that , with the operator defined in Eq. 4, to write out the Lagrangian L in the canonical coordinates as (see also Eq. 3) The neuronal dynamics is derived from requiring a stationary action (see Eq. 5), which is generally solved by the Euler-Lagrange equations (see Eq. 6). Because ũ only arises in L in the compound , the derivative of L with respect to ũ is identical to the derivative with respect to , Using the lookahead operator Eq. 14, the Euler-Lagrange equations can then be rewritten as Since and , the derivative of L with respect to ũ is the same as the derivative of L with respect to . Plugging this into Eq. 19, the Euler-Lagrange equations become a function of u and , Notice that, if we would have directly calculated , the second-order time derivative of the discounted future voltage would be absorbed in a first-order time derivative of the voltage. The reason is that , and only arises in this combination because the Lagrangian L = L(u) is only a function of u and not of . Hence, the acceleration term disappears, while a voltage derivative appears.

The solution of this differential equation Eq. 20 is , and hence any trajectory which satisfy the Euler-Lagrange equations will hence cause to converge to zero with a characteristic time scale of τ. Since we require that the initialisation is at t0 =−∞, we conclude that , as required in the rt-DeEP Theorem.

Deriving the network dynamics from the Euler-Lagrange equations

We now derive the equations of motion from the Euler-Lagrange equations. Noticing that u enters in twice, directly and through , and once in the output error ē*, we calculate from Eq. 16, using and W = (Win, Wnet), Remember that for non-output neurons i no target exist, and for those we set . Next, we apply the lookahead operator to this expression, as required by the Euler-Lagrange equations Eq. 19. In general , and we set for the expression on the right-hand side of Eq. 21, , which at the same time is . Hence, the Euler-Lagrange equations in the form of Eq. 20, , translate into To move from the middle to the last equality we replaced e by . In the last equality we interpret e as the sum of the two errors, e = ϵ + βe*, again using the middle equality. This proves Eq. 7.

Notice that the differential equation in Eq. 22 represents an implicit ordinary differential equation as on the right-hand side not only u, but also appears (in r and e). The uniqueness of the solution u(t) for a given initial condition is only guaranteed if it can be converted into an explicit ordinary differential equation (see Sect. C).

In taking the temporal derivative we assumed small learning rates such that terms including can be neglected. The derived dynamics for the membrane potential of a neuron ui in Eq. 22 show the usual leaky behavior of biological neurons. However, both presynaptic rates and prediction errors ēi enter the equation of motion with lookaheads, i.e., they are advanced and , cancelling the low-pass filtering. Since , the rate and error, ri and ei, can also be seen as nonlinear extrapolations from the voltage and its derivative into the future.

The instantaneous transmission of information throughout the network at the level of the voltages can now be seen by low-pass filtering Eq. 22 with initialization far back in the past, with column vector and . Hence, solving the voltage dynamics for u (Eq. 7a), with apical voltage derived from Eq. 7b, yields the somatic voltage u satisfying the self-consistency equation (Eq. 23) at any time. In other words, u and ē ‘propagate instantaneously’.

Deriving the error backpropagation formula

For clarity, we derive the error backpropagation algorithm for layered networks here. These can be seen as a special case of a general network with membrane potentials u and all-to-all weight matrix W (as introduced in appendix I), where the membrane potentials decompose into layerwise membrane potential vectors ul and the weight matrix into according block diagonal matrices Wl (with Wl being the weights that project into layer l).

Assuming a network with N layers, by low-pass filtering the equations of motion we get for all l ∈ 1, .., N, with the output error ēo of the general recurrent network becoming the error in the last layer, that itself is the target error, . The error , that we obtain from the general dynamics with , see Eq. 21 and Eq. 22, translates to an iterative formula for the error at the current layer l given the error at the downstream layer l+1, inherited from the drive of that downstream layer via Wl+1, and ēN = β ē* for the output layer. The learning rule that reduces ēl by gradient descent is proportional to this error and the presynaptic rate, as stated by Theorem 1, is for l = 1…N. Eq. 25 and Eq. 26 together take the form of the error backpropagation algorithm, where an output error is iteratively propagated through the network and used to adjust the weights in order to reduce the output cost C. From this, it is easy to see that without output nudging (i.e., β = 0), the output error vanishes and consequently all other prediction errors vanish as well, for all l ≤N. This also means that in the absence of nudging, no weight updates are performed by the plasticity rule.

The learning rule for arbitrary connectivities is obtained in the same way by dropping the layer-wise notation. In this case, low-pass filtering the equations of motion yields , as calculated in Eq. 23, and the low-pass filtered error , as inferred from Eqs 21 and 22. Hence, the plasticity rule in general reads

Proving Theorem 1 (rt-DeEP)

The implicit assumption in Theorem 1 is that exists in the distributional sense for t >−∞, which is the case for delta-functions in rin and step-functions in u*. Both parts (i) and (ii) of the Theorem are based on the requirement of stationary action δA = 0, and hence on u satisfying the Euler-Lagrange equations in the form of Eq. 22, . From the solution we conclude that for initialization at t0 = −∞ we have for all t. It is the latter stronger condition that we require in the proof. With this, the main ingredient of the proof follows is the mathematical argument of Scellier & Bengio, 2017, according to which the total and partial derivative of L with respect to W are identical, and this in our case is true for any time t, For convenience we considered to be a column vector, deviating from the standard notations (see tutorial end of sec:Integration). Analogously to Eq. 28, we infer . Reading Eq. 28 from the right to the left, we conclude that the learning rule for all β > 0 is gradient descent on L, i.e. . This total derivative of L can be analyzed for large and small β.

We show that in the limit of large β, becomes gradient descent on the mismatch energy. For this we first show that there is a solution of the self-consistency equation that is uniformly bounded for all t and β. For this we assume that the transfer function ρ(u) is non-negative, monotonically increasing and bounded, that its derivative ρ′(u) is bounded too, and that the input rates rin and the target potentials are also uniformly bounded. To show that under these conditions we always find an uniformly bounded solution u(t), we first consider the case where the output voltages are clamped to the target, such that ē* = 0. For simplicity we assume that ρ′(u) = 0 for |u| ≥ c0. For voltages u with ui ≤ c0 the recurrent input current is bounded, say for some c1 > c0. When including the error term , the total current still remains uniformly bounded, say |F (u)j |≤ c2 for all u with ui ≤ c0. Because for larger voltages ui > c0 the error term vanishes due to a vanishing derivative ρ′ (ui) = 0, the mapping F (u) maps the c2-box u (for which |ui| ≤ c2) onto itself. Brouwer’s fixed point theorem then tells us that there is a fixed point u = F (u) within the c2-box. The theorem requires the continuity of F, and this is assured if the neuronal transfer function is continuous.

We next relax the voltages of the output neurons from their clamped stage, . Remember that these voltages satisfy at any time t. We determine the correction term such that in the limit β → ∞ we get . The correction remains finite, and in the limit must be equal to . For arbitrary large nudging strengt h β, the output voltage uo deviates arbitrary little from the target voltage, , with target error shrinking like c2/β. Likewise, also for non-output neurons i, the self-consistency solution ui = F (u)i deviates arbitrarily little from the solution of the clamped state. To ensure the smooth drift of the fixed point while 1/β deviates from 0 we require that the Jacobian of F at the fixed point is invertible.

Because the output shrinks with 1/β, the cost shrinks quadratically with increasing nudging strength, , and hence the cost term that enters in vanishes in the limit β → ∞. In this large β limit, where and hence the outputs are clamped, , the Lagrangian reduces to the mismatch energy, L = EM. Along the least-action trajectories we therefore get . The first equality uses Eq. 28, and the second uses L = EM just derived for β = ∞. This is statement (i) of Theorem 1. In the case of successful learning, EM = 0, we also conclude that the cost vanishes, C = 0. This is the case because EM = 0 implies EM = 0 for all output neurons o. Since , we conclude that ēo = 0, and if the output neurons do not feed back to the network (which we can assume without loss of generality), we conclude that .

To consider the case of small β, we use that the cost C can be expressed as . This is a direct consequence of how C enters in , see Eq. 16 and Scellier & Bengio, 2017. We now put this together with Eq. 28 and the finding that . Since for the Lipschitz continuous function L in u, W and β (L is even smooth in these arguments), the total derivatives interchange (which is a consequence of the Moore-Osgood theorem applied to the limits of the difference quotients), we then get at any t,

The last expression is calculated from the specific form of the Lagrangian (Eq. 17), using that by definition . Finally, in the absence of output nudging, β = 0, we can assume vanishing errors, ē = 0, as they solve the self-consistency equation, for all t, see Eq. 27. For these solutions we have . Writing out the total derivative of the function with respect to β at β = 0 as limit of the difference quotient, , using that g(0) = 0, we calculate at any t, Here we assume that is evaluated at β > 0 (that itself approaches 0), while is evaluated at β = 0.Combining Eq. 29 and 30 yields the cost gradient at any t, This justifies the gradient learning rule in Eq. 27. Learning is stochastic gradient descent on the expected cost, where stochasticity enters in the randomization of the stimulus and target sequences rin(t) and u*(t). For the regularity statement, see ‘From implicit to explicit differential equations’ in the sec:Integration. Notice that this proof works for a very general form of the Lagrangian L, until the specific expression for . For a proof in terms of partial derivatives only, see SI, Sect. I, where also a primer on partial and total derivatives is found (SI, Sect. H).

Instantaneous gradient descent on

The cost at each time t is a function of the voltage uo of the output neurons and the corresponding targets. In a feedforward network, due to the instantaneity of the voltage propagation (Eq. 23), uo is in the absence of output nudging (β = 0) an instantaneous function of the voltage at the first layer, . For initialisation at t = −∞, the second term vanishes for all t and hence . The output voltage uo(t) therefore becomes a function FW of the low-pass filtered input rate rin(t) that captures the instantaneous network mapping, , and with this the cost also becomes an instantaneous function of and , namely .

For a general network, again assuming t0 = −∞, the voltage is determined by the vanishing gradient with , see Eq. 21. For the inclusive treatment of the initial transient see SI, Sects C and D. Remember that and . For a given and at time t, the equation f (u, t) = 0 can be locally solved for u if the Hessian is invertible, . This mapping can be restricted to the output voltages uo on the left-hand side, while replacing in the argument on the right-hand side (even if this again introduces uo there). With this we obtain the instantaneous mapping from the low-pass filtered input and the output error to the output itself. Notice that for functional feedforward network, the network weight matrix Wnet is lower triangular, and for small enough β the Hessian H is therefore always positive definite (see also Methods, Sect. F).

Proving Theorem 2 (rt-DeEL)

Here we restrict ourselves to layered network architectures. To prove Theorem 2 first assume that interneurons receive no nudging (βI = 0) and only the lateral interneuron-to-pyramidal weights are plastic. This is already sufficient to prove the rt-DeEL theorem. Yet, simulations showed that learning the lateral pyramidal-to-interneuron weights via top-down nudging, so that the interneuron activity mimics the upper layer pyramidal neuron activity, helps in learning a correct error representation. We consider this case of learning later.

If the microcircuits is ought to correctly implement error backpropagation, all local prediction errors ēl must vanish in the absence of output nudging (β = 0) as there is no target error. Consequently, any remaining errors in the network are caused by a misalignment of the lateral microcircuit. We show how learning the interneuron-to-pyramidal weights corrects for such misalignments.

To define the gradient descent plasticity of the weights from the interneurons to the pyramidal neurons, we consider the apical error formed by the difference of top-down input and interneuron input, , and define the apical mismatch energy as . Gradient descent along this energy with respect to yields evaluated online while presenting input patterns from the data distribution to the network. We assume that the apical contribution to the somatic voltage is further modulated by the somatic spike rate, . After successful learning, the top-down input Blul+1 is fully subtracted away by the lateral input in the apical compartment, and we have Once this condition is reached, the network achieves a state where, over the activity space spanned by the data, top-down prediction errors throughout the network vanish, We show that this top-down prediction error, after the s uccessful learning of the microcircuit, shares the properties of error-backpropagation for a suitable backprojection weights B.

Due to the vanishing prediction errors, pyramidal cells only receive bottom-up input . Using this expression as well as the expression for interneuron membrane potentials without top-down nudging (βI = 0 in Eq. 9), , and plugging both into Eq. 33, we get Assuming that has full rank, and the low-pass filtered rates span the full nl dimensions of layer l when sampled across the data set, we conclude that In other words, the loop via upper layer and back is learned to be matched by a lateral loop through the interneurons. Eq. 36 imposes a restriction on the minimal number of interneurons at layer l. In fact, the matrix product BlWl+1 maps a nl-dimensional space onto itself via nl+1-dimensional space. The maximal rank of the this matrix product is limited by the smallest dimension, i.e. rank(BlWl+1) ≤ min(nl, nl+1). Analogously, rank . But since the two ranks are the same according to Eq. 36, we conclude that in general must hold, i.e. there should be at least as many interneurons at layer l as the lowest number of pyramidal neurons at either layer l or l+1. Note that by choosing as in (Sacramento et al., 2018) (or as in this work), the conditions is fulfilled.

With and Eq. 36, the top-down prediction error from Eq. 34, in the presence of output nudging (β > 0), can be written in the backpropagation form Finally, the simulations showed that learning the lateral weights in the microcircuit greatly benefits from also adapting the pyramidal-to-interneuron weights W IP by gradient descent on , using top-down nudging of the inhibitory neurons (βI > 0), After learning we have , and plugging in (Eq. 9), we obtain . Since , we conclude as before, The top-down weights that nudge the lower-layer interneurons has randomized entries and may be considered as full rank. If there are less pyramidal neurons in the upper layer than interneurons in the lower layer, selects a subspace in the interneuron space of dimension . This seems to simplify the learning of the interneuron-to-pyramidal cell connections . In fact, this learning now has only to match the nl+1-dimensional interneuron subspace embedded in dimensions to an equal (nl+1-)dimensional pyramidal cell subspace emedded in nl dimensions.

Learning of the interneuron-to-pyramidal cell connections works with the interneuron nudging as before, and combining Eqs 36 with 39 yields the ‘loop consistency’ The learning of the microcircuit was described in the absence of output nudging. Conceptually, this is not a problem as one could introduce a pre-learning phase where the lateral connections are first correctly aligned before learning of the feedforward weights begins. In simulations we find that both the lateral connections as well as the forward connections can be trained simultaneously, without the need for such a pre-learning phase. We conjecture that this is due to the fact that our plasticity rules are gradient descent on the energy functions L, EPI and EIP respectively.

From implicit to explicit differential equations

The voltage dynamics is solved by a forward-Euler scheme . The derivative is calculated either through (i) the implicit differential equation (Eq. 7) yielding , or (ii) by isolating and solving for the explicit differential equation , as explained in SI, Sect. C (after Eq. 51).

The implicit differential equation, , see Eq. 22, is iteratively solved by assigning and calculating the error with and .

This iteration exponentially converges to a fixed point on a time scale , where 1 − k > 0 is the smallest Eigenvalue of the Hessian , see SI, Sect. C.

The explicit differential equation is obtained by eliminating the from the right-hand side of the implicit differential equation. Since enters linearly we get with . The explicit form is obtained by matrix inversion, , as the Hessian is invertible if it is strictly positive definite (which is typically the case, see SI, Sect. C, after Eq. 48). The external input and the target enter through , where the derivative of the target voltage is only added for the output neurons o. This explicit differential equation is shown to be contractive in the sense that for each input trajectory rin(t) and target trajectory u*(t), the voltage trajectory u(t) is locally attracting for neighbouring trajectories. This local attracting trajectory is the vanishing-gradient trajectory f (u, t) = 0, and the gradient remains 0 even if the input contains delta-functions, see SI Sect. D.

Moving and latent equilibria: a formal definition

We showed that the motor output (uo), together with the low-pass filtered sensory input and the motor feedback is in a moving equilibrium, , see Fig. 3a. In general, a dynamical system in u that is given in an implicit form with external inputs is said to be in a moving equilibrium if the variable u is an instantaneous function of the input x at any point in time, u = F (x). The fact that the implicit differential equation G = 0 represents a dynamical system in u implies that, in principle, it has a representation in the explicit form , guaranteed by an invertible Jacobian .

Our example is obtained from with and , leading to . Given the paramterization of G by the weights, we get the parametrized function u = FW (x), and this is restricted to the output components u of u. The condition on the Jacobian translates to being invertible. Crucially, the description of the dynamics in the biological or physical substrate is not given in its explicit form . However, it is given in an implicit form expressed as , where still appears on the right-hand side. This ‘hybrid’ form is directly solved either in real time by the biophysical substrate itself, or by the forward-Euler scheme on clocked hardware, see (i) above. Notice that moving equilibria u = FW (x) with are able to capture complex temporal processing of the instantaneous input rin. In fact, the low-pass filtering can be obtained on various time scales through different τin’s, and FW for a general network W can be arbitrary complex. The task is to adapt W such that the ‘hybrid’ dynamical system eventually implements the target mapping .

The Latent Equilibrium (Haider et al., 2021) can be analogously formalized as a dynamical system in u, implicitly given by , and having a solution of the form . Abbreviating again with the same Lagrangian as in the present NLA, the Latent Equilibrium is obtained for . The solution implies that the rate is an instantaneous function of , here without low-pass filtering. As for moving equilibria, the crucial point is that the biophysical substrate implements a hybrid form of the dynamical system, now , that is implicitly solved by the analog substrate, and also allows for a solution in clocked hardware. For an extended stability analysis of the moving and latent equilibria see SI, Sect. D.

Simulation details

Solving the explicit differential equation seems to be more robust when the learning rate for gets larger. The explicit form is also less sensitive to large Euler steps dt, see SI Sect. C. By this reason, the ordinary differential equations (ODE) were solved in the explicit form when including plasticity . The algorithms are summarized as follows, once without interneurons (Algo 1), and once with interneurons (Algo 2):

with projection neurons only, for Figs 3 & 4 (using the explicit ODE, i.e. Step 12 instead of 11)

including plastic interneurons, for Fig. 5 (using the explicit ODE, i.e. Step 13 instead of 12)

Details for Fig. 3b

Color coded snapshot of cortical local field potentials (LFPs) in a human brain from 56 deep iEEG electrodes at various locations, converted with the sigmoidal voltage-to-rate function and plotted onto a standard Talairach Brain (Talairach & Tournoux, 1988). The iEEG data is from a patient with pharmacoresistant epilepsy and electrodes implanted during presurgical evaluation, extracted from the data release of Burrello et al., 2019. The locations of the electrodes are chosen in accordance with plausibilty, as the original positions of the electrodes were omitted due to ethical standards to prevent patient identification.

Details for Fig. 3c

Simulations of the voltage dynamics (Eq. 7a) and weight dynamics (Eq. 8), with learning rate η = 10−3, step size dt = 1 ms for the forward Euler integration, membrane time constant τ = 10 ms and logistic activation function. Weights were initialized randomly from a normal distribution 𝒩 (0, 0.12) with a cut-off at ±0.3. The number of neurons in the network 𝒩 was n = 96, among them 56 output neurons 𝒪 ⊂ 𝒩 that were simultaneously nudged, and 40 hidden neurons. During training, all output neurons were nudged simultaneously (with β = 0.1), whereas during testing, only 42 out of 56 neurons were nudged, the remaining 14 left to reproduce the traces. Data points of the iEEG signal were sampled with a frequency of 512Hz. For simplicity, we therefore assumed that successive data points are separated by 2ms, and up-sampled the signal via simple interpolation to 1ms resolution as required by our integration scheme. Furthermore, the raw values were normalized by dividing them by a factor of 200 to ensure that they are approximately in a range of ±1 −2. Training and testing was done on two separate 8s traces of the iEEG recording. Same data as in Fig. 3b1.

Details for Fig. 4

Simulation of the neuronal and synaptic dynamics as given by Eq. 7a, Eq. 7b and Eq. 8. For 5 ms, 10 ms and 50 ms presentation time, we used an integration step size of dt = 0.05 ms, dt = 0.1 ms and dt = 0.5 ms, respectively (and dt = 1 ms otherwise). As an activation function, we used the step-linear function (hard sigmoidal) with for for u ≥ 1 and in between. The learning rate was initially set to η = 10−3 and then reduced to η = 10−4 after 22 000 s. The nudging strength was β = 0.1 and the membrane time constant τ = 10 ms. In these simulations (and only for these) we assumed that at each presynaptic layer l = 0, 1, .., n −1 there is a first neuron indexed by 0 that fires with constant rate , effectively allowing the postsynaptic neurons to learn a bias through the first column of the weight matrix Wl+1. Weights were initialized randomly from a normal distribution 𝒩 (0, 0.012) with a cut-off at ±0.03. For an algorithmic conversion see the scheme below. In Fig. 4c1, ‘rt-DeEP w/o lookahead’ is based on the dynamics . For ‘ w/o error + backprop’, we use as the forward model (so without error terms on the membrane potential, but a prospective r), and calculate weight updates using error backpropagation. In Fig. 4c2, we provide three controls: the test error for (i) a standard shallow artificial neural network trained on MNIST (black dashed line), (ii) rt-DeEP without prospective coding (as in Fig. 4c1), but in c2 with plasticity only turned on when the network is completely stationary, i.e., after waiting for several 100ms, such that synaptic weights are not changed during transients (orange dashed line, denoted by ‘w/o transients’), and (iii) an equivalent artificial neural network, , trained using error backpropagation (black dashed line, ‘standard backprop’).

Details for Fig. 5

Simulation of neuronal and synaptic dynamics with plastic microcircuit, i.e., the pyramidal-to-interneuron and lateral weights of the microcircuit learned during training.

For the results shown in Fig. 5c2, the following parameters were used. As an activation function, we used a hard sigmoid function and the membrane time constant was set to τ = 10ms. Image presentation time is 100 ms. Forward, pyramidal-to-interneuron and interneuron-to-pyramidal weights were initialized randomly from a normal distribution 𝒩 (0, 0.012) with a cut-off at ±0.03. All learning rates were chosen equal η = 10−3 and were subsequently reduced to η = 10−4 after 22000s training time. The nudging parameters were set to β = 0.1 and . The feedback connections Bl and the nudging matrices were initialized randomly from a normal distribution 5 𝒩 (0, 0.012) with a cut-off at ±0.15. The used integration step size was dt = 0.25ms. All weights were trained simultaneously. For an algorithmic conversion see the scheme below. The interneuron membrane potential was calculated by Eq. 9 with a linear transfer function.